---
title: "GWAS catalog traits"
author: Gibran Hemani
date: "2023-08-25"
categories: []
---

## Background

Identify GWAS catalog traits to import to OpenGWAS

```{r}
library(dplyr)
library(tidyr)
library(data.table)
library(ieugwasr)
library(ggplot2)
```

Get list of harmonised studies in GWAS catalog

```{r}
download.file("ftp://ftp.ebi.ac.uk/pub/databases/gwas/summary_statistics/harmonised_list.txt", "harmonised_list.txt")
a <- scan("harmonised_list.txt", what=character())
head(a)
```

Get EBI IDs already in OpenGWAS

```{r}
ao <- ieugwasr::gwasinfo()
gotebi <- grep("ebi-a", ao$id, value=TRUE) %>% gsub("ebi-a-", "", .)
```

Identify datasets not available

```{r}
a <- tibble(path=a) %>%
    tidyr::separate(path, sep="/", into=c("dot", "range", "id", "harmonised", "fn"), remove=FALSE) %>% 
    as_tibble %>%
    tidyr::separate(fn, sep="-", into=c("pmid", "id2", "efo"), remove=FALSE) %>%
    filter(!is.na(id2)) %>%
    filter(!id %in% gotebi) %>%
    dplyr::select(-c(id2, dot)) %>%
    mutate(path=gsub("^./", "", path))
a
```

There are 4427 that have no pmid, remove for now.

Summarise how many studies by pmid:

```{r}
as <- a %>% group_by(pmid) %>% summarise(n=n()) %>% arrange(desc(n))
as %>% head(n=20)
```

How many traits per study?

```{r}
qplot(y=as$n, x=1:nrow(as)) +
scale_y_log10() +
labs(y="Number of traits in study", x="Study")
```

How many traits to import if we only keep studies with fewer than 1000 traits

```{r}
subset(as, n <= 1000) %>% {sum(.$n)}
```

How many traits to import if we only keep studies with fewer than 200 traits

```{r}
subset(as, n <= 200) %>% {sum(.$n)}
```

Studies to ignore (for now at least)

```{r}
pmid_ignore <- c(
    34662886, # rare variant aggregate enrichment of ukb exome data
    35078996, # pqtl
    34737426, # ukb analysis (duplication of existing work)
    35668104, # lipids superseded by others
    33983923  # facial variation - 1k traits
)
```


```{r}
a_keep <- subset(a, !pmid %in% pmid_ignore)
write.table(a_keep$path, file="gwascat_keeplist.txt", row=F, col=F, qu=F)
```

---

```{r}
sessionInfo()
```
