---
title: "Difference in difference models in observational data are still potentially confounded"
author: "Gibran Hemani"
date: "2022-01-09"
---

Motivated by discussion with Tim Cadman relating to this [Renzi et al (2019). Long-Term PM10 Exposure and Cause-Specific Mortality in the Latium Region (Italy): A Difference-in-Differences Approach](https://ehp.niehs.nih.gov/doi/full/10.1289/EHP3759).

For argument's sake suppose that air pollution (PM10) isn't causal for deaths. Simulate a situation where wealth varies by region, and each region has a different wealth trajectory over time. Some info:

-   Wealth, PM10 and deaths are measured for 10 years in 300 regions.
-   Wealth causes PM10 and deaths.
-   There is a global confounder for the start point of wealth and deaths
-   Time has an additional effect on both deaths and wealth - i.e. it's also a global confounder
-   The causal effect of wealth on death = 1. We want our model to get that right.

```{r}
library(dplyr)
nregion <- 300
region <- 1:300
wealth_int <- rnorm(nregion)
wealth_slope <- rnorm(nregion)
global_confounder <- rnorm(nregion, sd=10)
nyear <- 10

dat <- lapply(1:nregion, function(i)
  {
    tibble(
      region=i,
      year=1:nyear,
      # wealth goes up due to global confounder, year, random error
      wealth = global_confounder[i] + wealth_int[i] + year * wealth_slope[i] + rnorm(nyear),
      # PM10 only related to wealth and random error
      pm10 = wealth + rnorm(nyear, sd=5),
      # deaths go up due to global confounders, wealth, year and random term
      deaths = global_confounder[i] + wealth + rnorm(nyear) + year
    )
}) %>%
  bind_rows()
```

```{r}
dat
```

Plot it showing change in deaths over time by region

```{r}
library(ggplot2)
ggplot(dat, aes(x=year, y=deaths)) +
  geom_point(aes(group=as.factor(region))) +
  geom_line(aes(group=as.factor(region)))
```

Try again but just regression lines per region

```{r}
ggplot(dat, aes(x=year, y=deaths)) +
  geom_point(aes(group=as.factor(region))) +
  geom_smooth(method="lm", aes(group=as.factor(region)), se=FALSE)
```

Show that PM10 is closely coupled with wealth

```{r}
plot(pm10 ~ wealth, dat)
```

Use regression to test for influence of wealth on deaths - this gives a very confounded result because of the global confounder.

```{r}
summary(lm(deaths ~ wealth, dat))
```

Same will be true for PM10

```{r}
summary(lm(deaths ~ pm10, dat))
```

So PM10 is confounded at two levels - the between-region (global) level and the within-region level.

Now do a DiD model for wealth - it should give us an unbiased estimate of 1. Note that this long form of DiD is sometimes called a fixed-effects model.

```{r}
summary(lm(deaths ~ wealth + as.factor(region) + as.factor(year), dat)) %>%
  coefficients %>% as_tibble %>% slice(n=2)
```

It does. What about DiD model for PM10? This should be less confounded because it eliminates the global confounder, but still confounded by the structural confounding that happens at all areas

```{r}
summary(lm(deaths ~ pm10 + as.factor(region) + as.factor(year), dat)) %>%
  coefficients %>% as_tibble %>% slice(n=2)
```

If we control for wealth, the effect of PM10 will be unbiased because the within-region bias has been removed, and the global bias acted via wealth anyway so that's been removed also

```{r}
summary(lm(deaths ~ pm10 + wealth, dat)) %>%
  coefficients %>% as_tibble %>% slice(n=2)
```

We could try to simplify by doing the more explicit difference in difference estimate. Compare the change in deaths with the change in wealth (or PM10) over the 10 year period.

```{r}
dat2 <- group_by(dat, region) %>%
  summarise(
    delta_deaths = deaths[nyear] - deaths[1],
    delta_wealth = wealth[nyear] - wealth[1],
    delta_pm10 = pm10[nyear] - pm10[1]
  )
dat2

```

Do the DiD estimates using these - should recapitulate what we got above

```{r}
summary(lm(delta_deaths ~ delta_wealth, dat2))
```

```{r}
summary(lm(delta_deaths ~ delta_pm10, dat2))
```

Weirdly, it works for wealth but we do get a slightly difference answer for PM10. There is probably a lot of literature on what makes a fixed effects model (the first version of the DiD we did above) different from an explicit version like this one.

To summarise - the DiD model is useful to account for unmeasured global confounders (including time), but it might not be too surprising that it doesn't control for all confounders - you really do need some sort of experiment / randomisation that specifically mimics the exact intervention you want to make to get to completely unconfounded effects.
