---
title: "Differential susceptibility explained by interactors and mediators with different distributions"
author: Gibran Hemani
date: "2023-06-09"
categories: [interactions, multi-ancestry]
---

## Background

What leads to different effect estimates in different populations? If the DAG is the same, then different distributions of mediators and interactions could explain different effects

```{r}
n <- 100000

g1 <- rbinom(n, 2, 0.3)
g2 <- rbinom(n, 2, 0.3)

# The liability for the mediator 
# Influenced by g (same effect in both populations)
# Influenced by some other variable that gives it a different mean in the two populations
# Now liability of mediator has different means
ml1 <- plogis(scale(g1) + rnorm(n, 2, 1))
ml2 <- plogis(scale(g2) + rnorm(n, 0, 1))

hist(ml1)
hist(ml2)
```

```{r}
# Generate mediators - they have different means
m1 <- rbinom(n, 1, ml1)
m2 <- rbinom(n, 1, ml2)
mean(m1)
mean(m2)
```

Disease influenced by mediator and noise (same effects)

```{r}
d1 <- rbinom(n, 1, plogis(m1 + rnorm(n)))
d2 <- rbinom(n, 1, plogis(m2 + rnorm(n)))

mean(d1)
mean(d2)
```

Slightly different prevalence in different populations. 

The genotypic effect is now different in the two populations due to different

```{r}
glm(d1 ~ g1, family="binomial")
glm(d2 ~ g2, family="binomial")
```



---

```{r}
sessionInfo()
```
