---
title: "Meta analysis including small samples"
author: Gibran Hemani
date: "2024-09-28"
categories: []
---

## Background

Does meta analysis with very small samples lead to bias?

- Simulate a regression
- Split the data into different size groups and meta analyse the groups
- Compare the meta-analysis results to the full analysis
- Does smaller group size lead to bias?

```{r}
library(dplyr)
library(ggplot2)
library(tidyr)
ivw <- function(betas, ses) {
  weights <- 1 / ses^2
  beta <- sum(weights * betas) / sum(weights)
  se <- 1 / sqrt(sum(weights))
  return(c(beta, se))
}
```


```{r}

fast_assoc <- function(y, x) {
	index <- is.finite(y) & is.finite(x)
	n <- sum(index)
	y <- y[index]
	x <- x[index]
	vx <- var(x)
	bhat <- stats::cov(y, x) / vx
	ahat <- mean(y) - bhat * mean(x)
	# fitted <- ahat + x * bhat
	# residuals <- y - fitted
	# SSR <- sum((residuals - mean(residuals))^2)
	# SSF <- sum((fitted - mean(fitted))^2)

	rsq <- (bhat * vx)^2 / (vx * var(y))
	fval <- rsq * (n-2) / (1-rsq)
	tval <- sqrt(fval)
	se <- abs(bhat / tval)

	# Fval <- (SSF) / (SSR/(n-2))
	# pval <- pf(Fval, 1, n-2, lowe=F)
	# p <- stats::pf(fval, 1, n-2, lower.tail=FALSE)
	return(list(
		ahat=ahat, bhat=bhat, se=se, fval=fval, n=n
	))
}

sim <- function(n, nbreak, beta, af) {
    g <- rbinom(n, 2, af)
    y <- g * beta + rnorm(n)
    full <- summary(lm(y ~ g))
    breaks <- cut(1:n, breaks=nbreak)
    r <- lapply(levels(breaks), \(l) {
        y <- y[breaks == l]
        g <- g[breaks == l]
        if(length(unique(g)) == 1) return(NULL)
        a <- fast_assoc(y, g)
        tibble(
            beta = a$bhat,
            se = a$se
        )
    }) %>% bind_rows()
    meta <- ivw(r$beta, r$se)
    tibble(
        n=n, nbreak=nbreak, beta=beta, af=af,
        bhat=c(full$coefficients[2, 1], meta[1]),
        se=c(full$coefficients[2, 2], meta[2]),
        what=c("full", "meta")
    )
}
sim(10000, 1000, 0.2, 0.5)

```


```{r}
param <- expand.grid(
    n = 10000,
    nbreak = c(10, 100, 1000),
    beta = c(0, 0.2),
    af = c(0.1, 0.5),
    nsim = 1:30
)
dim(param)
```

```{r}
res <- lapply(1:nrow(param), \(i) {
    sim(param$n[i], param$nbreak[i], param$beta[i], param$af[i]) %>% mutate(sim=param$nsim[i])
}) %>% bind_rows()
```

```{r}
ggplot(res, aes(x=as.factor(nbreak), y=bhat, color=what)) +
    geom_boxplot() +
    facet_grid(beta ~ af, labeller=label_both)
```

```{r}
resw <- pivot_wider(res, names_from=what, values_from=c(bhat, se))
ggplot(resw, aes(x=bhat_full, y=bhat_meta)) +
geom_point() +
facet_wrap(~beta + af + nbreak, scale="free", labeller=label_both) +
geom_abline(slope=1, intercept=0)
```


## Try again with just continuous data

```{r}
sim2 <- function(n, nbreak, beta) {
    g <- rnorm(n)
    y <- g * beta + rnorm(n)
    full <- summary(lm(y ~ g))
    breaks <- cut(1:n, breaks=nbreak)
    r <- lapply(levels(breaks), \(l) {
        y <- y[breaks == l]
        g <- g[breaks == l]
        if(length(unique(g)) == 1) return(NULL)
        a <- fast_assoc(y, g)
        tibble(
            beta = a$bhat,
            se = a$se
        )
    }) %>% bind_rows()
    meta <- ivw(r$beta, r$se)
    tibble(
        n=n, nbreak=nbreak, beta=beta,
        bhat=c(full$coefficients[2, 1], meta[1]),
        se=c(full$coefficients[2, 2], meta[2]),
        what=c("full", "meta")
    )
}
sim2(10000, 1000, 0.2)
sim2(10000, 100, 0.2)
sim2(10000, 10, 0.2)
```

```{r}
param2 <- expand.grid(
    n = 10000,
    nbreak = c(10, 100, 1000),
    beta = c(0, 0.2),
    nsim = 1:100
)
dim(param2)
```

```{r}
res2 <- lapply(1:nrow(param2), \(i) {
    sim2(param2$n[i], param2$nbreak[i], param2$beta[i]) %>% mutate(sim=param2$nsim[i])
}) %>% bind_rows()
```

```{r}
resw <- pivot_wider(res2, names_from=what, values_from=c(bhat, se))
ggplot(resw, aes(x=bhat_full, y=bhat_meta)) +
geom_point() +
facet_wrap(~beta + nbreak, scale="free", labeller=label_both) +
geom_abline(slope=1, intercept=0) +
geom_smooth(method="lm")
```

```{r}
ggplot(resw, aes(x=se_full, y=se_meta)) +
geom_point() +
facet_wrap(~beta + nbreak, scale="free", labeller=label_both) +
geom_abline(slope=1, intercept=0) +
geom_smooth(method="lm")
```

```{r}
res2 %>% 
    mutate(pval = pnorm(abs(bhat)/se, low=F)) %>%
    filter(beta == 0) %>%
    group_by(nbreak, what) %>%
    summarise(power = sum(pval < 0.05)/n(), n=n())
```


## Summary

- No bias
- The effects are not identical and become more noisy as the sample size decreases. 
- The standard error is under estimated for the meta analysis as the sample sizes decrease, which leads to slight elevation of type 1 error. 

---

```{r}
sessionInfo()
```
