---
title: "Meta regression using exposure effect size"
author: Gibran Hemani
date: "2023-07-09"
categories: []
---

## Background

1. Simulate individual level data in system in which u causes x and y, and both x and u have independent instruments
2. Identify instruments (should include many gx and some gu)
3. Estimate heterogeneity contribution from each instrument
4. Meta regression of instrument strength against heterogeneity


```{r}
library(simulateGP)
library(TwoSampleMR)

# 1. simulate system
nid <- 100000
nsnp <- 1000
gx <- make_geno(nid, nsnp, 0.4)
gu <- make_geno(nid, nsnp, 0.4)
bx <- choose_effects(nsnp, sqrt(0.4))
bu <- choose_effects(nsnp, sqrt(0.4))
u <- make_phen(bu, gu)
x <- make_phen(c(bx, 0.4), cbind(gx, u))
y <- make_phen(c(0.4, 0.4), cbind(x, u))
```

```{r}
# 2. Identify instruments
dat <- get_effs(x, y, cbind(gx, gu))
dats <- subset(dat, pval.exposure < (0.05/2000))
str(dat)
```

```{r}
table(dats$SNP > 1000)
```

```{r}
dats$strength <- cut(-log10(dats$pval.exposure), breaks=quantile(-log10(dats$pval.exposure), probs=seq(0, 1, 0.1)))
table(dats$strength)
```

```{r}
# Overall MR estimate
res <- mr(dats, method="mr_ivw")
res
```

Stratify by instrument strength

```{r}
dats %>% group_by(strength) %>% do({mr(., method="mr_ivw")}) %>% 
    ggplot(., aes(x=strength, y=b)) +
    geom_point() +
    geom_errorbar(aes(ymin=b-se*1.96, ymax=b+se*1.96), width=0) +
    geom_hline(yintercept=0.4) +
    labs(x="Instrument strength", y="MR effect estimate")
```

```{r}
# 3. Estimate heterogeneity
ss <- mr_singlesnp(dats) %>% filter(!grepl("All", SNP)) %>% mutate(SNP = as.numeric(SNP))
ss$qj <- (1/ss$se^2) * (res$b - ss$b)^2
ss$str <- (dats$beta.exposure/dats$se.exposure)^2
ss$str2 <- -log10(dats$pval.exposure)
mr_heterogeneity(dats, method="mr_ivw")$Q == sum(ss$qj)
```

4. Meta regression

```{r}
# All SNPs
summary(lm(qj ~ str, data=ss))
```

Higher strength means lower heterogeneity

```{r}
# Only x snps
summary(lm(qj ~ str, data=subset(ss, SNP <= 1000)))
```

Wouldn't expect this - weaker SNPs should contribute more heterogeneity due to weak instrument bias

```{r}
# Only u snps
summary(lm(qj ~ str, data=subset(ss, SNP > 1000)))
```


```{r}
plot(qj ~ str, ss)
```


```{r}
plot(qj ~ str, ss %>% filter(SNP > 1000))
```



```{r}
plot(qj ~ str, ss %>% filter(SNP <= 1000))
```

Use all SNPs in X only to look at heterogeneity wrt instrument strength

```{r}
ssx <- mr_singlesnp(dat[1:1000,]) %>% filter(!grepl("All", SNP)) 
ssx$qj <- (1/ssx$se^2) * (res$b - ssx$b)^2
ssx$str <- (dat[1:1000,]$beta.exposure/dat[1:1000,]$se.exposure)^2

summary(lm(qj ~ str, ssx))

ggplot(ssx, aes(x=str, y=qj)) +
geom_point() +
geom_smooth()
```

```{r}
datx <- subset(dat, SNP %in% (1:1000))
datx$strength <- cut(-log10(datx$pval.exposure), breaks=quantile(-log10(datx$pval.exposure), probs=seq(0, 1, 0.1)))
table(datx$strength)
datx %>% group_by(strength) %>% do({mr(., method="mr_ivw")}) %>% 
    ggplot(., aes(x=strength, y=b)) +
    geom_point() +
    geom_errorbar(aes(ymin=b-se*1.96, ymax=b+se*1.96), width=0) +
    geom_hline(yintercept=0.4) +
    labs(x="Instrument strength", y="MR effect estimate")
```

No impact of weak instrument bias on heterogeneity...?

## Empirical example

```{r}
meta_reg_dat <- function(d){
    d <- subset(d, mr_keep)
    res <- mr(d, method="mr_ivw")
    bivw <- subset(res, method == "Inverse variance weighted")$b

    ss <- mr_singlesnp(d) %>% filter(!grepl("All", SNP))
    ss$qj <- (1/ss$se^2) * (bivw - ss$b)^2
    ss$str <-  (d$beta.exposure/d$se.exposure)^2
    return(ss)
}

strat_inst_mr <- function(d){
    d$fval <- (d$beta.exposure/d$se.exposure)^2
    d$strength <- cut(d$fval, breaks=quantile(d$fval, probs=seq(0, 1, 0.1)))
    d %>% group_by(strength) %>% do({mr(., method="mr_ivw")}) %>% 
        ggplot(., aes(x=strength, y=b)) +
        geom_point() +
        geom_errorbar(aes(ymin=b-se*1.96, ymax=b+se*1.96), width=0) +
        geom_hline(yintercept=0.4) +
        labs(x="Instrument strength", y="MR effect estimate")

}

d <- make_dat("ukb-b-10787", "ebi-a-GCST007431")
dfev <- make_dat("ukb-b-10787", "ukb-b-19657", prox=F)

ssrat <- meta_reg_dat(d)
head(ssrat)
summary(lm(qj ~ str, ssrat))
```

```{r}
strat_inst_mr(dfev)
```

```{r}
ggplot(ssrat, aes(x=log(str), y=qj)) +
geom_point() +
geom_smooth()
```




## Summary

- U instruments do contribute more to heterogeneity
- However, there is quite a strong relationship between instrument strength and heterogeneity. I would have expected lower strength to have more weak instrument bias and therefore more heterogeneity, however it is actually the opposite - higher instrument strength has higher heterogeneity amongst X SNPs
- There is a very strong relationship between instrument strength and heterogeneity amongst U SNPs, this makes sense as the average effect is more weighted towards X SNPs and so stronger U SNPs create a larger departure

---

```{r}
sessionInfo()
```
