---
title: "Decomposing drug X side effect matrices"
author: Gibran Hemani
date: "2023-10-31"
categories: []
---

## Background

**Aim:** Want to predict side effects using MR. 

Have a matrix of side effects x drugs. 

```
A = n_se X m_drugs
```

Each drug binds some genes

```
B = p_genes X m_drugs
```

MR of each gene on all traits

```
C = q_traits X p_genes
```

and a matrix linking trait terms to side effect terms

```
D = q_traits X n_se
```

## Basic simulation

- m=3 drugs
- n=5 side effects
- p=6 genes
- q=10 traits

```{r}
# gene x drug - e.g. based on binding affinities
B <- matrix(c(
    0, 1, 1,
    1, 0, 0,
    0, 1, 0,
    0, 0, 1,
    0, 0, 0,
    0, 1, 1
), 6, 3)

# trait x se - matches trait names to side effect terms
D <- matrix(c(
    1, 0, 0, 0, 0,
    1, 0, 0, 0, 0,
    0, 1, 0, 0, 0,
    0, 1, 0, 0, 0,
    0, 0, 1, 0, 0, 
    0, 0, 1, 0, 0,
    0, 0, 1, 0, 0,
    0, 0, 1, 0, 0,
    0, 0, 0, 1, 0,
    0, 0, 0, 1, 0
), 10, 5)

# True mapping of genes to side effects - we don't observe this
gse <- matrix(c(
    1, 0, 0, 0, 0, 0,
    1, 1, 0, 0, 0, 0,
    0, 0, 1, 0, 0, 0,
    0, 0, 0, 1, 1, 0,
    0, 0, 0, 0, 1, 0
), 5, 6)

# True drug x side effect matrix is generated from gene side effects by gene drug binding
A <- gse %*% B
A
```


We don't actually see the gse matrix. If everything works as we hypothesise then the trait x gene matrix that we observe would follow:

```{r}
C <- D %*% gse
C
```

Now we have B, C and D. How do we get back to A? Need to invert D, which isn't square so use Moore-Penrose pseudoinverse

```{r}
library(pracma)
Ahat <- pinv(D) %*% C %*% B
```

Does the prediction match the true A?

```{r}
cor(c(Ahat), c(A))
```

```{r}
plot(Ahat, A)
```

Quite close - the pseudoinverse has failed to get some of the values correct. Alternative to using pseudoinverse is to just manually re-label trait names with side effect values

---

```{r}
sessionInfo()
```
