---
title: "Estimating environmental factor using GxE interaction"
author: Gibran Hemani
date: "2025-01-21"
categories: []
---

## Background

Is it possible to estimate the environmental factor using the GxE interaction term? In this model the genetic effect is masked unless individuals experience a particular environment. Stratifying individuals by different levels of environmental exposure will give different marginal genetic effects. Can the influence of the environment be recovered from this?

## Initial simulation

```{r}
library(purrr)
library(ggplot2)
library(dplyr)
```


```{r}
fn <- function(n, bap, bm, bi, sim=1) {
    args <- list(n=n, bap=bap, bm=bm, bi=bi, sim=sim) %>% as_tibble()
    ap <- rbinom(n, 2, 0.5) / 2
    g <- rbinom(n, 2, 0.3)

    l <- ap * bap + g * ap * bi + rnorm(n)
    copd <- rbinom(n, 1, plogis(l))
    bhat <- glm(copd ~ g * ap, family="binomial")$coef[4]
    args$bhat <- bhat
    return(args)
}
fn(100000, 0.3, 0.1, 0.5)
```


```{r}
param <- expand.grid(
    n=100000,
    bap=c(0.1, 0.3, 0.5),
    bm=c(0, 0.1),
    bi=c(0.1, 0.3, 0.5),
    sim=1:5
)
res <- pmap(param, fn)
```

```{r}
res %>% 
    bind_rows() %>%
    ggplot(aes(bap, bhat, color=factor(bi))) +
    geom_point()
```


Untidy sims:

```{r, eval=FALSE}

library(dplyr)
library(simulateGP)

n <- 10000
G <- correlated_binomial(n, 0.3, 0.1, 0.5)
dim(G)
cor(G)

lm(G[,1] ~ G[,2])

x <- G[,1] + rnorm(n)

lm(x ~ G[,2])

cor(G) * (sd(G[,1]) / sd(G[,2]))







n <- 100000
ap <- rbinom(n, 2, 0.5) / 2
g <- rbinom(n, 2, 0.3)

b <- 1.2

l <- ap * b + g + g * ap + rnorm(n)
copd <- rbinom(n, 1, plogis(l))

b0 <- glm(copd ~ g, family="binomial", subset=ap==0)$coef[2]
b1 <- glm(copd ~ g, family="binomial", subset=ap==1)$coef[2]
b2 <- glm(copd ~ g, family="binomial", subset=ap==2)$coef[2]

b0
b1
b2

b2 <- glm(copd ~ g * ap, family="binomial")


fn <- function(n, bap, bm, bi, sim=1) {
    args <- list(n=n, bap=bap, bm=bm, bi=bi, sim=sim) %>% as_tibble()
    ap <- rbinom(n, 2, 0.5) / 2
    g <- rbinom(n, 2, 0.3)

    l <- ap * bap + g * ap * bi + rnorm(n)
    copd <- rbinom(n, 1, plogis(l))
    bhat <- glm(copd ~ g * ap, family="binomial")$coef[4]
    args$bhat <- bhat
    return(args)
}

param <- expand.grid(
    n=100000,
    bap=c(0.1, 0.3, 0.5),
    bm=c(0, 0.1),
    bi=c(0.1, 0.3, 0.5),
    sim=1:5
)

fn(100000, 0.3, 0.1, 0.5)

library(purrr)
res <- pmap(param, fn)


library(ggplot2)

res %>% 
    bind_rows() %>%
    ggplot(aes(bap, bhat, color=factor(bi))) +
    geom_point()


```


---

```{r}
sessionInfo()
```
