---
title: "MR and imperfect clumping"
author: Gibran Hemani
date: "2023-07-13"
categories: []
---

## Background

4000 instruments for educational attainment using clumping r2 = 0.1, and se doubles when using r2 = 0.001.

That smaller standard error is either due to the R2 in the exposure being higher or the non-independence of effects artificially increasing precision, or a mixture of both.

So the question is the impact of the latter – if we have some true correlation structure with realistic F stats at a specific locus, and then we try to clump at r2 = 0.001 vs 0.1, how many instruments do we retain (it should be 1) and if more than 1, what is that impact on the standard error


```{r}
library(dplyr)
library(simulateGP)
library(TwoSampleMR)
library(purrr)
library(ggplot2)
```

Simulate causal snp g + another that is correlated with it

```{r}
n <- 100000
g <- correlated_binomial(n, p1=0.5, p2=0.5, rho=sqrt(0.1))
# x caused by just snp 1
x <- g[,1] * 0.5 + rnorm(n)
y <- x * 0.5 + rnorm(n)

# MR using both SNPs, treating as if they are independent
get_effs(x, y, g) %>% mr(method="mr_ivw") %>% str
```

```{r}
# MR using just the causal SNP
get_effs(x, y, g[,1, drop=F]) %>% mr(method=c("mr_ivw", "mr_wald_ratio")) %>% str
```

There's hardly any difference in the SE here. Try over a range of scenarios

```{r}
param <- expand.grid(
    r2=seq(0, 1, by=0.02),
    bgx=seq(0,1, by=0.2),
    bxy=seq(0,1, by=0.2),
    n=100000
)
param$sim <- 1:nrow(param)
dim(param)

res <- map(1:nrow(param), \(i){
    g <- correlated_binomial(param$n[i], p1=0.5, p2=0.5, rho=sqrt(param$r2[i]))
    # x caused by just snp 1
    x <- g[,1] * param$bgx[i] + rnorm(n)
    y <- x * param$bxy[i] + rnorm(n)

    bind_rows(
        get_effs(x, y, g) %>% {suppressMessages(mr(., method="mr_ivw"))},
        get_effs(x, y, g[,1, drop=F]) %>% {suppressMessages(mr(., method="mr_wald_ratio"))}
    ) %>% mutate(sim=param$sim[i]) %>% return()
}) %>% bind_rows %>% inner_join(param, ., by="sim")
```

Standard errors across all scenarios

```{r}

ggplot(res, aes(x=r2, y=se)) +
geom_point(aes(colour=as.factor(nsnp))) +
geom_line(aes(colour=as.factor(nsnp))) +
facet_grid(bgx ~ bxy, labeller=label_both, scale="free_y")
```

Bias across all scenarios:

```{r}
ggplot(res, aes(x=r2, y=b)) +
geom_point(aes(colour=as.factor(nsnp))) +
geom_line(aes(colour=as.factor(nsnp))) +
facet_grid(bgx ~ bxy, labeller=label_both)
```

Look at just one

```{r}
ggplot(res %>% filter(bgx == 0.2, bxy == 0.2), aes(x=r2, y=se)) +
geom_point(aes(colour=as.factor(nsnp))) +
geom_line(aes(colour=as.factor(nsnp))) +
facet_grid(bgx ~ bxy, labeller=label_both, scale="free_y")
```


How much does SE reduce if you increase the number of instruments

```{r}
n <- 100000
g <- correlated_binomial(n, p1=0.5, p2=0.5, rho=sqrt(0.1))
# x caused by just snp 1
x <- g %*% c(0.5, 0.5) + rnorm(n)
y <- x * 0.5 + rnorm(n)

# MR using both SNPs, treating as if they are independent
get_effs(x, y, g) %>% mr(method="mr_ivw") %>% str
```

```{r}
get_effs(x, y, g[,1, drop=F]) %>% mr(method=c("mr_ivw", "mr_wald_ratio")) %>% str
```

## Summary

- Relaxed r2 e.g. from 0 to 0.1 doesn't seem to have a huge impact on standard errors
- In the one SNP situation relaxed r2 has no impact on bias, and could only plausibly change things under substantial heterogeneity which correlates with overrepresentation.
- More realistic simulations would look at whether this changes when the p-value at the second locus is very large, and would also look at the probability of erroneously keeping multiple loci for a single causal variant
- Some instability in SEs when correlated SNPs used
- Increasing SNPs will reduce SE but need to know change in r2 to examine this properly

---

```{r}
sessionInfo()
```
