---
title: "Conditioning on the exposure"
author: Gibran Hemani
date: "2026-06-22"
categories: []
---

## Background

A tempting solution to the pleiotropy problem in MR is to check if the instrument associates with the outcome conditioning on the exposure. Under the assumption that X and Y are confounded, this opens a collider path to Y. Meaning that by default we expect the instrument to associate with the outcome conditional on the exposure. This is a problem because it means that we cannot use this as a test for pleiotropy.

This simulation examines how sensitive the association between the instrument and outcome is to conditioning on the exposure. The simulation is based on a simple model where X and Y are confounded by U, and the instrument Z is associated with X but not Y. i.e. There is no pleiotropy - so any association between Z and Y conditional on X is due to the collider path, and will lead to erroneously disgarding the instrument.

## Simulation

```{r}
library(tidyverse)
library(furrr)
library(here)

fast_assoc <- function(y, x)
{
	index <- is.finite(y) & is.finite(x)
	n <- sum(index)
	y <- y[index]
	x <- x[index]
	vx <- var(x)
	bhat <- stats::cov(y, x) / vx
	ahat <- mean(y) - bhat * mean(x)
	# fitted <- ahat + x * bhat
	# residuals <- y - fitted
	# SSR <- sum((residuals - mean(residuals))^2)
	# SSF <- sum((fitted - mean(fitted))^2)

	rsq <- (bhat * vx)^2 / (vx * var(y))
	fval <- rsq * (n-2) / (1-rsq)
	tval <- sqrt(fval)
	se <- abs(bhat / tval)

	# Fval <- (SSF) / (SSR/(n-2))
	# pval <- pf(Fval, 1, n-2, lowe=F)
	p <- stats::pf(fval, 1, n-2, lower.tail=FALSE)
	return(list(
		ahat=ahat, bhat=bhat, se=se, fval=fval, pval=p, n=n
	))
}

make_dat <- function(n=1000, bxy=0.5, bxz=0.5, byu=0.5, bxu=0.5)
{
    u <- rnorm(n)
    z <- rnorm(n)
    x <- bxz * z + bxu * u + rnorm(n)
    y <- bxy * x + byu * u + rnorm(n)

    dat <- data.frame(u=u, z=z, x=x, y=y)
    return(dat)
}

estimation <- function(dat)
{
    # estimate the association between Z and Y
    res1 <- fast_assoc(dat$y, dat$z)
    res2 <- fast_assoc(dat$x, dat$z)

    # estimate the association between Z and Y conditional on X
    res2 <- fast_assoc(lm(y ~ x, data=dat)$residuals, dat$z)

    tibble(
        rsq_xy=cor(dat$x, dat$y)^2,
        rsq_xz=cor(dat$x, dat$z)^2,
        rsq_yz=cor(dat$y, dat$z)^2,
        rsq_yu=cor(dat$y, dat$u)^2,
        rsq_xu=cor(dat$x, dat$u)^2,
        pval_xz=res1$pval,
        fval=res1$fval,
        pval=res1$pval,
        pval_cond=res2$pval
    )
}

sim <- function(n=1000, bxy=0.5, bxz=0.5, byu=0.5, bxu=0.5)
{
    dat <- make_dat(n=n, bxy=bxy, bxz=bxz, byu=byu, bxu=bxu)
    res <- estimation(dat)
    return(res)
}

sim(n=100000, bxy=0.5, bxz=0.2, byu=0.5, bxu=0.5)

```


```{r, eval=FALSE}
params <- expand.grid(
    n=c(100000),
    bxy=c(0),
    bxz=seq(0, 0.04, by=0.01),
    byu=seq(0, 1, by=0.2),
    bxu=seq(0, 1, by=0.2),
    simrep=1:100
)

dim(params)
plan(multisession, workers=8)
results <- future_pmap(
    params,
    function(n, bxy, bxz, byu, bxu, simrep) {
        sim(n=n, bxy=bxy, bxz=bxz, byu=byu, bxu=bxu)
    }, .options = furrr_options(seed = TRUE), .progress = TRUE
) %>% bind_rows()

# results <- params[1:100,] |>
#   rowwise() |>
#   mutate(sim = list(sim(n=n, bxy=bxy, bxz=bxz, byu=byu, bxu=bxu))) |>
#   unnest(sim)

results <- bind_cols(results, params)
saveRDS(results, here("posts/2026-06-22-conditioning-ont-he-exposure", "results.rds"))
```


```{r}
results <- readRDS(here("posts/2026-06-22-conditioning-ont-he-exposure", "results.rds"))
sumres <- results %>%
    group_by(bxu) %>% mutate(rsq_xu = mean(rsq_xu) %>% round(2)) %>%
    group_by(byu) %>% mutate(rsq_yu = mean(rsq_yu) %>% round(2)) %>%
    ungroup() %>%
    group_by(bxz, rsq_yu, rsq_xu) %>%
    summarise(
        mean_rsq_xz = mean(rsq_xz),
        mean_pval_xz = mean(-log10(pval_xz)),
        mean_pval_cond = mean(pval_cond),
        pow_cond = mean(pval_cond < 0.05)
    ) %>%
    ungroup()


ggplot(sumres, aes(x=mean_rsq_xz, y=pow_cond)) +
  geom_point() +
  geom_line() +
  labs(x="SNP-exposure R2", y="Power for Z-Y association conditional on X") +
  geom_hline(yintercept=0.05, linetype="dashed", alpha=0.5) +
  facet_grid(rsq_yu ~ rsq_xu, labeller = label_both)

```

---

```{r}
sessionInfo()
```
