---
title: "Conditioning on the exposure"
author: Gibran Hemani
date: "2026-06-22"
categories: []
---

## Background

A tempting solution to the pleiotropy problem in MR is to check if the instrument associates with the outcome conditioning on the exposure. Under the assumption that X and Y are confounded, this opens a collider path to Y. Meaning that by default we expect the instrument to associate with the outcome conditional on the exposure. This is a problem because it means that we cannot use this as a test for pleiotropy.

This simulation examines how sensitive the association between the instrument and outcome is to conditioning on the exposure. The simulation is based on a simple model where X and Y are confounded by U, and the instrument Z is associated with X but not Y. i.e. There is no pleiotropy - so any association between Z and Y conditional on X is due to the collider path, and will lead to erroneously disgarding the instrument.

## Simulation

```{r}
library(tidyverse)
library(furrr)
library(here)

fast_assoc <- function(y, x)
{
	index <- is.finite(y) & is.finite(x)
	n <- sum(index)
	y <- y[index]
	x <- x[index]
	vx <- var(x)
	bhat <- stats::cov(y, x) / vx
	ahat <- mean(y) - bhat * mean(x)
	# fitted <- ahat + x * bhat
	# residuals <- y - fitted
	# SSR <- sum((residuals - mean(residuals))^2)
	# SSF <- sum((fitted - mean(fitted))^2)

	rsq <- (bhat * vx)^2 / (vx * var(y))
	fval <- rsq * (n-2) / (1-rsq)
	tval <- sqrt(fval)
	se <- abs(bhat / tval)

	# Fval <- (SSF) / (SSR/(n-2))
	# pval <- pf(Fval, 1, n-2, lowe=F)
	p <- stats::pf(fval, 1, n-2, lower.tail=FALSE)
	return(list(
		ahat=ahat, bhat=bhat, se=se, fval=fval, pval=p, n=n
	))
}

make_dat <- function(n=1000, bxy=0.5, bxz=0.5, byu=0.5, bxu=0.5)
{
    u <- rnorm(n)
    z <- rnorm(n)
    x <- bxz * z + bxu * u + rnorm(n)
    y <- bxy * x + byu * u + rnorm(n)

    dat <- data.frame(u=u, z=z, x=x, y=y)
    return(dat)
}

estimation <- function(dat)
{
    # estimate the association between Z and Y
    res1 <- fast_assoc(dat$y, dat$z)
    res2 <- fast_assoc(dat$x, dat$z)

    # estimate the association between Z and Y conditional on X
    res2 <- fast_assoc(lm(y ~ x, data=dat)$residuals, dat$z)

    tibble(
        rsq_xy=cor(dat$x, dat$y)^2,
        rsq_xz=cor(dat$x, dat$z)^2,
        rsq_yz=cor(dat$y, dat$z)^2,
        rsq_yu=cor(dat$y, dat$u)^2,
        rsq_xu=cor(dat$x, dat$u)^2,
        pval_xz=res1$pval,
        fval=res1$fval,
        pval=res1$pval,
        pval_cond=res2$pval
    )
}

sim <- function(n=1000, bxy=0.5, bxz=0.5, byu=0.5, bxu=0.5)
{
    dat <- make_dat(n=n, bxy=bxy, bxz=bxz, byu=byu, bxu=bxu)
    res <- estimation(dat)
    return(res)
}

sim(n=100000, bxy=0.5, bxz=0.2, byu=0.5, bxu=0.5)

```


```{r, eval=FALSE}
params <- expand.grid(
    n=c(100000),
    bxy=c(0),
    bxz=seq(0, 0.04, by=0.01),
    byu=seq(0, 1, by=0.2),
    bxu=seq(0, 1, by=0.2),
    simrep=1:100
)

dim(params)
plan(multisession, workers=8)
results <- future_pmap(
    params,
    function(n, bxy, bxz, byu, bxu, simrep) {
        sim(n=n, bxy=bxy, bxz=bxz, byu=byu, bxu=bxu)
    }, .options = furrr_options(seed = TRUE), .progress = TRUE
) %>% bind_rows()

# results <- params[1:100,] |>
#   rowwise() |>
#   mutate(sim = list(sim(n=n, bxy=bxy, bxz=bxz, byu=byu, bxu=bxu))) |>
#   unnest(sim)

results <- bind_cols(results, params)
saveRDS(results, here("posts/2026-06-22-conditioning-ont-he-exposure", "results.rds"))
```


```{r}
results <- readRDS(here("posts/2026-06-22-conditioning-ont-he-exposure", "results.rds"))
sumres <- results %>%
    group_by(bxu) %>% mutate(rsq_xu = mean(rsq_xu) %>% round(2)) %>%
    group_by(byu) %>% mutate(rsq_yu = mean(rsq_yu) %>% round(2)) %>%
    ungroup() %>%
    group_by(bxz, rsq_yu, rsq_xu) %>%
    summarise(
        mean_rsq_xz = mean(rsq_xz),
        mean_pval_xz = mean(-log10(pval_xz)),
        mean_pval_cond = mean(pval_cond),
        pow_cond = mean(pval_cond < 0.05)
    ) %>%
    ungroup()


ggplot(sumres, aes(x=mean_rsq_xz, y=pow_cond)) +
  geom_point() +
  geom_line() +
  labs(x="SNP-exposure R2", y="Power for Z-Y association conditional on X") +
  geom_hline(yintercept=0.05, linetype="dashed", alpha=0.5) +
  facet_grid(rsq_yu ~ rsq_xu, labeller = label_both)

```


## What if we have lots of instruments?

Is there a constant collider bias, meaning that all residual effects of instrument on Y will be the same + horizontal pleiotropy?

```{r}

gwas <- function(z, y) {
    for(i in 1:ncol(z)) {
        Z <- as.numeric(unlist(z[,i]))
        res <- fast_assoc(y, Z)
        if (i == 1) {
            out <- as_tibble(res) %>% mutate(snp=i)
        } else {
            out <- bind_rows(out, as_tibble(res) %>% mutate(snp=i))
        }
    }
    return(out)
}

make_dat <- function(params)
{
    nsnp <- params$nsnp
    prop_pleio <- params$prop_pleio
    n <- params$n
    bxy <- params$bxy
    bxz <- params$bxz
    byu <- params$byu
    bxu <- params$bxu
    bzx <- params$bzx
    bzy <- params$bzy

    u <- rnorm(n)
    z <- matrix(rnorm(n * nsnp), nrow=n, ncol=nsnp)
    zx <- z %*% bzx %>% drop
    zy <- z %*% bzy %>% drop
    x <- bxz * zx + bxu * u + rnorm(n)
    y <- bxy * x + zy + byu * u + rnorm(n)

    z <- as.data.frame(z)
    names(z) <- paste0("z", 1:nsnp)

    dat <- tibble(u=u, x=x, y=y) %>% bind_cols(z)
    return(dat)
}


make_params <- function(nsnp=100, prop_pleio=0.5, n=1000, bxy=0.5, bxz=0.5, byu=0.5, bxu=0.5) {
    bzx <- rnorm(nsnp, 0, 0.1)
    bzy <- rnorm(nsnp, 0, 0.1) * sample(c(0,1), nsnp, replace=TRUE, prob=c(1-prop_pleio, prop_pleio))

    list(
        nsnp=nsnp,
        prop_pleio=prop_pleio,
        n=n,
        bxy=bxy,
        bxz=bxz,
        byu=byu,
        bxu=bxu,
        bzx=bzx,
        bzy=bzy
    )
}

estimation <- function(dat)
{
    # estimate the association between Z and Y
    # res1 <- gwas(dat[,4:ncol(dat)], dat$y)
    # res2 <- gwas(dat[,4:ncol(dat)], dat$x)

    # estimate the association between Z and Y conditional on X
    res3 <- gwas(dat[,4:ncol(dat)], lm(y ~ x, data=dat)$residuals)
    return(res3)
    bind_rows(
        res1 %>% mutate(type="Z-Y"),
        res2 %>% mutate(type="Z-X"),
        res3 %>% mutate(type="Z-Y|X")
    )
}
```


```{r}
p <- make_params(n=100000)
dat <- make_dat(p)
res <- estimation(dat)
plot(res$bhat ~ p$bzy)
```

```{r}
p <- make_params(n=100000, byu = 1, bxu = 1, bxy=0.5, prop_pleio = 0.5)
dat <- make_dat(p)
res <- estimation(dat)
plot(I(res$bhat) ~ p$bzy)
plot(I(res$bhat) ~ p$bzx)
summary(lm(I(res$bhat/p$bzx) ~ p$bzy))
```

```{r}
lapply(1:10, \(i) {
    p <- make_params(n=100000, nsnp=1, byu=)
})
```



Expected collider bias = buy * bux / (bux^2 + var(x)) * bgx

```{r}
sim <- function(buy=0.1, bux=0.1, vx=4, bgx=1) {
    n <- 100000
    u <- rnorm(n)
    g <- rnorm(n)
    gpred <- bgx * g
    uxpred <- bux * u
    resid <- rnorm(n, 0, sqrt(vx - var(gpred)) - var(uxpred))
    x <- gpred + uxpred + resid
    y <- buy * u + rnorm(n)
    cond <- residuals(lm(y ~ x))
    bhat <- lm(cond ~ g)$coef[2]
    exp <- -buy * bux / (bux^2 + var(resid)) * bgx
    tibble(
        buy=buy, bux=bux, vx=vx, bgx=bgx,
        varx=var(x), varu=var(u), varg=var(g),
        exp=exp, obs=bhat
    )
}

param <- expand.grid(
    buy=seq(0, 1, by=0.1),
    bux=seq(0, 1, by=0.1),
    vx=c(4),
    bgx=c(0.5, 1)
)

theoretical_res <- lapply(1:nrow(param), \(i) {
    sim(param$buy[i], param$bux[i], param$vx[i], param$bgx[i])
}) %>% bind_rows()

ggplot(theoretical_res, aes(x=exp, y=obs)) +
    geom_point() +
    geom_abline(slope=1, intercept=0, linetype="dashed", alpha=0.5) +
    facet_grid(bgx ~ vx, labeller = label_both) +
    labs(x="Expected collider bias", y="Observed collider bias")

```


In MR there is a question of how much horizontal
  pleiotropy is going to bias the IV estimate. I am starting to have the opinion
  that when you have a large number of SNPs, there are two classes of pleiotropy
  to be concerned about. 1) The SNP impacts Y through multiple independent
  paths, i.e. it is biologically complex; 2) The SNP more influences a confounder
  of X and Y, and arises as an instrument for Y.

  One way to estimate the extent of (1) happening, would be to test if SNPs
  associate with Y conditional on X. A big problem with this approach is that X
  and Y are typically confounded, so conditioning on X will induce an association
  between the SNP and Y through the backdoor path via the confounder. But this
  bias would be a constant bias * the bgx effect, so it could be possible to (a)
  determine the extent of confounding and (b) determine the extent of deviation
  of all SNPs from the expected collider effect.

---

```{r}
sessionInfo()
```
