---
title: "Concentration index"
author: Gibran Hemani
date: "2025-03-21"
categories: []
---

## Background

How do concentration index and Lorenz curves respond to different simulation scenarios?


## Simulations

### Random

GWAS attention is independent of DALY

```{r}
library(dplyr)
library(rineq)

ndisease <- 450
diseases <- tibble(
    disease = 1:ndisease, 
    daly = rbeta(ndisease, 1, 0.7), 
    gwas_attention = rbeta(ndisease, 1, 0.7))
ci1 <- ci(ineqvar=diseases$gwas_attention, outcome=diseases$daly, method="direct")
summary(ci1)
plot(ci1)
```

Try again with uniform distribtuion

```{r}
diseases <- tibble(
    disease = 1:ndisease, 
    daly = rbeta(ndisease, 1, 1), 
    gwas_attention = rbeta(ndisease, 1, 1))
ci1 <- ci(ineqvar=diseases$gwas_attention, outcome=diseases$daly, method="direct")
summary(ci1)
plot(ci1)
```

### Proportional

GWAS attention is the same as DALY

```{r}
ndisease <- 450
diseases <- tibble(
    disease = 1:ndisease, 
    daly = rbeta(ndisease, 1, 0.7), 
    gwas_attention = daly + rnorm(ndisease, 0, sd=0.001))
ci1 <- ci(ineqvar=diseases$gwas_attention, outcome=diseases$daly, method="direct")
summary(ci1)
plot(ci1)
```

Again with uniform distribution

```{r}
diseases <- tibble(
    disease = 1:ndisease, 
    daly = rbeta(ndisease, 1, 1), 
    gwas_attention = daly + rnorm(ndisease, 0, sd=0.001))
ci1 <- ci(ineqvar=diseases$gwas_attention, outcome=diseases$daly, method="direct")
summary(ci1)
plot(ci1)
```


```{r}
diseases <- tibble(
    disease = 1:ndisease, 
    gwas_attention = rbeta(ndisease, 1, 1), 
    daly = gwas_attention + rnorm(ndisease, 0, sd=1))
ci1 <- ci(ineqvar=diseases$gwas_attention, outcome=diseases$daly, method="direct")
summary(ci1)
plot(ci1)
```


### Unequal

GWAS attention grows faster than DALY

```{r}
diseases <- tibble(
    disease = 1:ndisease, 
    daly = 1:ndisease, 
    gwas_attention = daly^2)
ci1 <- ci(ineqvar=diseases$gwas_attention, outcome=diseases$daly, method="direct")
summary(ci1)
plot(ci1)
```

Note that the x-axis is ranked so it loses information about the non-uniformity of the distribution

```{r}
diseases <- tibble(
    disease = 1:ndisease, 
    daly = 1:ndisease, 
    gwas_attention = daly)
ci1 <- ci(ineqvar=diseases$gwas_attention, outcome=diseases$daly, method="direct")
summary(ci1)
plot(ci1)
```



```{r}
diseases <- tibble(
    disease = 1:ndisease, 
    daly = (1:ndisease)^2, 
    gwas_attention = 1:ndisease)
ci1 <- ci(ineqvar=diseases$gwas_attention, outcome=diseases$daly, method="direct")
summary(ci1)
plot(ci1)
```

---

```{r}
sessionInfo()
```




