---
title: "Gene-environment equivalence"
author: Gibran Hemani
date: "2023-03-15"
categories: []
---

## Background

Latest PRS for height explains \~40% of the variance. Heritability \~0.7. Height has a negative association with coronary heart disease. Height is fixed after adolescence, which means it can only be influenced by a) genetic factors; b) early life exposures; c) stochasticity.

Question: Is the influence of height on CHD different when the modifier is genetic vs non-genetic?

Analysis: Estimate the association of height PRS on CHD, and then residualise height for the PRS and determine the residual height association with CHD.

## Model

### Gene environment equivalence

Under gene-environment equivalence, the following model:

$$
y_i = \alpha + \beta_{xy} x_i + \epsilon_i
$$

where $y_i$ is the outcome (e.g. CHD), $\beta_{xy}$ is the causal effect of the exposure on the outcome and

$$
x_i = a + \sqrt{h^2} g_i + \sqrt{1-h^2} e_i
$$

where $x_i$ is the exposure (height) in the $i$th individual, $h^2$ is the heritability of height, $g_i$ is the (perfectly measured) genetic value for individual $i$ and $e_i$ is the non-genetic component of height. There is no confounding in this simple model. **What is the expected association of g on y?**

$$
\begin{aligned}
\beta_{gy} &= \frac{cov(y, g)}{var(g)} \\
&= \frac{cov(\beta_{xy}x, g)}{ var(g) } \\
&= \frac{cov(\beta_{xy}\sqrt{h^2}g, g)}{ var(g) } \\
&= \frac{\beta_{xy}\sqrt{h^2}var(g)}{ var(g) } \\
&= \beta_{xy}\sqrt{h^2}
\end{aligned}
$$

So this is expected and the basic result for MR. **What is the expected association of x on y after it has been residualised for g?** First get the residual assuming perfect information of $g$

$$
\begin{aligned}
\hat{e_i} &= x_i - \hat{\sqrt{h^2}}g_i \\
&\approx \sqrt{1-h^2}e_i
\end{aligned}
$$

Now find the association of $\hat{e}_i$ with $y_i$

$$
\begin{aligned}
\beta_{\hat{e}y} &= \frac{cov(\hat{e}, y)}{var(\hat{e})} \\
&= \frac{cov(\sqrt{1-h^2}e, \beta_{xy}(\sqrt{1-h^2}e)}{(1-h^2)var(e)} \\
&= \frac{\beta_{xy}(1-h^2)var(e)}{(1-h^2)var(e)} \\
&= \beta_{xy}
\end{aligned}
$$

**RESULT:** The residual height association with y is expected to be equal to the raw height association with y.

### Gene environment non-equivalence

Now allow the genetic and non-genetic components of height to have independent influences on CHD.

$$
x_i = \sqrt{h^2}g_i + \sqrt{1-h^2}e_i
$$

as before, and now

$$
y_i = \beta_g g_i + \beta_e e_i + \epsilon_i
$$

Expected association of x on y

$$
\begin{aligned}
\beta_{xy} &= \frac{cov(\sqrt{h^2}g_i + \sqrt{1-h^2}e_i,\beta_g g_i + \beta_e e_i)}{var(x)} \\
&= \frac{\sqrt{h^2}\beta_g var(g) + \sqrt{1-h^2}\beta_e var(e)}{h^2 var(g) + (1-h^2) var(e)} \\
\end{aligned}
$$

Assuming $var(g)=var(e)=1$ this reduces to

$$
\beta_{xy} = \sqrt{h^2}\beta_g + \sqrt{1-h^2}\beta_e
$$

**Now what is the expected association of x on y after it has been residualised for g?**

$$
\begin{aligned}
\beta_{\hat{e}y} &= cov(\hat{e}, y) / var(\hat{e}) \\
&= \frac{cov(\sqrt{1-h^2}e, \beta_g g + \beta_e e)}{(1-h^2)var(e)} \\
&= \frac{\sqrt{1-h^2}\beta_evar(e)}{(1-h^2)var(e)} \\
&= \sqrt{\frac{(1-h^2)\beta_e^2 var(e)^2}{(1-h^2)^2 var(e)^2}} \\
&= \frac{\beta_e}{\sqrt{1-h^2}}
\end{aligned}
$$

## Check with simulations

### Gene environment equivalence

```{r}
library(dplyr)
library(ggplot2)
n <- 100000
bxy <- 0.3
h2 <- 0.7
g <- rnorm(n)
e <- rnorm(n)
x <- sqrt(h2) * g + sqrt(1-h2) * e
y <- x * bxy + rnorm(n)
```

Get residual

```{r}
xres <- residuals(lm(x ~ g)) # sim
xres_exp <- sqrt(1-h2) * e # expected
# check they're the same
cor(xres, xres_exp)
lm(xres ~ xres_exp)$coef[2]
```

Assoc of x on y should be

```{r}
lm(y ~ x)$coef[2] # sim
bxy # expected
```

Expected assoc of g on y

```{r}
lm(y ~ g)$coef[2] # sim
sqrt(h2)*bxy # expected
```

Expected assoc of residual x on y

```{r}
lm(y ~ xres)$coef[2] # sim
bxy
```

### Gene environment non-equivalence

```{r}
n <- 100000
bg <- 0.3
be <- 0.6
h2 <- 0.7
g <- rnorm(n)
e <- rnorm(n)
x <- sqrt(h2) * g + sqrt(1-h2) * e
y <- g * bg + e * be + rnorm(n)
```

Get residual

```{r}
xres <- residuals(lm(x ~ g)) # sim
xres_exp <- sqrt(1-h2) * e # expected
# check they're the same
cor(xres, xres_exp)
lm(xres ~ xres_exp)$coef[2]
```

Association of xres and y

```{r}
lm(y ~ xres)$coef[2] # sim
be / sqrt(1-h2)
```

Assoc of x and y - simulation

```{r}
lm(y ~ x)$coef[2] # sim
bg * sqrt(h2) + be * sqrt(1-h2) # exp
```

## Incomplete adjustment of g

Does incomplete adjustment of g change things much?

### Gene environment equivalence

```{r}
n <- 100000
bxy <- 0.3
h2 <- 0.7
g_exp <- 0.5 # proportion of g explained by prs
g_explained <- rnorm(n) * sqrt(g_exp)
g_unexplained <- rnorm(n) * sqrt(1-g_exp)
g <- g_explained + g_unexplained
e <- rnorm(n)
x <- sqrt(h2) * g + sqrt(1-h2) * e
y <- x * bxy + rnorm(n)
```

Get residual - now this is different from expected above due to residual including some unadjusted genetic variance, but linearly still related

```{r}
xres <- residuals(lm(x ~ g_explained)) # sim
xres_exp <- sqrt(1-h2) * e # expected
# check they're the same
cor(xres, xres_exp)
lm(xres ~ xres_exp)$coef[2]
```

What is the PRS (explained g) assoc with y?

```{r}
lm(y ~ g_explained)$coef[2]
```

Assoc of height with y

```{r}
lm(y ~ x)$coef[2]
```

And residual with y

```{r}
lm(y ~ xres)$coef[2]
```

**So the residual x effect remains identical to the raw effect of x.**

## In progress... (ignore for now)

Does confounding change things much?

```{r}
n <- 100000
bxy <- 0.3
bu <- 0.3
h2 <- 0.7
u <- rnorm(n)
g <- rnorm(n)
e <- rnorm(n)
x <- sqrt(h2) * g + sqrt(1-h2) * e + u * bu
y <- x * bxy + rnorm(n) + u * bu
```

y \~ x

```{r}
lm(y ~ x)$coef[2]
```

y \~ x_res

```{r}
xres <- residuals(lm(x ~ g)) # sim
xres_exp <- sqrt(1-h2) * e # expected
# check they're the same
cor(xres, xres_exp)
lm(xres ~ xres_exp)$coef[2]
```

```{r}
lm(y ~ xres)$coef[2]

```

```{r}
lm(y ~ xres_exp)$coef[2]
```

```{r}
lm(y ~ g)$coef[2]
```

```{r}
testdat <- function(y, x, g)
{
  xres <- residuals(lm(x ~ g))
  tribble(~model, ~beta,
          "y ~ x", lm(y ~ x)$coef[2],
          "y ~ g", lm(y ~ g)$coef[2],
          "y ~ xres", lm(y ~ xres)$coef[2]
  )
}
```

No confounding, gene environment equivalence

```{r}
n <- 100000
bxy <- -0.3
h2 <- 0.7
g <- rnorm(n)
e <- rnorm(n)
x <- sqrt(h2) * g + sqrt(1-h2) * e
y <- x * bxy + rnorm(n)
testdat(y, x, g)
```

```{r}
n <- 100000
buy <- 0.3
bux <- -0.3
bxy <- 0.3
h2 <- 0.7
u <- rnorm(n)
g <- rnorm(n)
e <- rnorm(n)
x <- sqrt(h2) * g + sqrt(1-h2) * e + bux * u
y <- x * bxy + rnorm(n) + buy * u
testdat(y, x, g)
```

```{r}
n <- 100000
buy <- 0.3
bux <- -0.3
bg <- -0.2
be <- -0.3
h2 <- 0.7
u <- rnorm(n)
g <- rnorm(n)
e <- rnorm(n)
x <- sqrt(h2) * g + sqrt(1-h2) * e + bux * u
y <- g * bg + e * be + rnorm(n) + buy * u
testdat(y, x, g)
```

```{r}
param <- expand.grid(
  buy = seq(-0.3, 0.3, by=0.3),
  bux = seq(-0.3, 0.3, by=0.3),
  bg = seq(-0.3, 0.3, by=0.3),
  be = seq(-0.3, 0.3, by=0.3)
)

res <- lapply(1:nrow(param), function(i) {
  n <- 100000
  buy <- param$buy[i]
  bux <- param$bux[i]
  bg <- param$bg[i]
  be <- param$be[i]
  h2 <- 0.7
  u <- rnorm(n)
  g <- rnorm(n, sd=sqrt(h2))
  e <- rnorm(n, sd=sqrt(1-h2))
  x <- g + e + bux * u
  y <- g * bg + e * be + rnorm(n) + buy * u
  testdat(y, x, g) %>% bind_cols(param[i,])
})
res <- bind_rows(res)
```

```{r}
res %>% mutate(ge = bg-be) %>%
  filter(bg == -0.3) %>%
  #filter(round(bg, 1) == -0.2, be < 0, round(buy, 1) %in% c(-0.3, 0, 0.3), round(bux, 1) %in% c(-0.3, 0, 0.3)) %>%
ggplot(., aes(x=beta, y=ge)) +
  geom_point(aes(colour=model)) +
  geom_line(aes(colour=model)) +
  facet_grid(buy ~ bux) +
  scale_colour_brewer(type="qual")
```

```{r}
n <- 100000
buy <- 0
bux <- 0
bg <- -0.4
be <- -0.2
h2 <- 0.5
u <- rnorm(n)
g <- rnorm(n, sd=sqrt(h2))
e <- rnorm(n, sd=sqrt(1-h2))
x <- g + e
y <- g * bg + e * be + rnorm(n)
testdat(y, x, g)
```

```{r}
library(simulateGP)
g <- make_geno(1000, 100, 0.3)
b <- rnorm(100)
prs <- g %*% b
ve <- var(prs) / 0.7 - var(prs)
e <- rnorm(1000, sd=sqrt(ve))
x <- prs + e
cor(x, prs)^2
var(prs) / var(x)
```
