---
title: "Power of G x age interactions"
author: Gibran Hemani
date: "2024-11-23"
categories: []
---

## Background

Individual level analysis versus age-stratified GWAS to detect GxAge interactions

## Simulation setup

```{r, message=FALSE, warning=FALSE}
library(dplyr)
library(metafor)
library(ggplot2)
# library(rms)
set.seed(12345)
```

The simulation is based on the following model:

$$
y_i = \alpha + \beta_0 g_i + \beta_1 g_i a_i + \beta_2 g_i a_i^2 + \beta_3 g_i a_i^3 + e_i
$$

where $y_i$ is the phenotype of individual $i$, $g_i$ is the genotype, $a_i$ is the age normalised to have range $(-1,1)$, and $e_i$ is the error term. 

The simulation is based on the following steps:

1. Generate a set of coefficients for the age polynomial
2. Generate a set of individuals with age, genotype, and error terms

```{r}
simdat <- function(n=500000, b0, b1, b2, b3) {
    nage <- length(40:70)
    coef <- tibble(
        age = 40:70,
        x = seq(-1, 1, length.out=nage),
        bxage = x^3 * b3 + x^2 * b2 + x * b1 + b0
    )

    dat <- tibble(
        age = runif(n, 40, 70) %>% round,
        g = rbinom(n, 2, 0.5),
        e = rnorm(n)
    ) %>% group_by(age) %>%
        do({
            a <- subset(coef, age == .$age[1])
            mutate(., prs = a$bxage * g, y = prs + e)
        })

    return(list(dat=dat, coef=coef))
}
```

An example of the simulation with **no interaction**:

```{r}
sim <- simdat(500000, 0.02, 0, 0, 0)
sim
```

Estimation is performed by 

1. generating summary statistics for 2-year age bins
2. estimating the effects of the genotype on the phenotype in each age bin using a simple linear model
3. performing a polynomial meta regression of 

$$
\beta_{age} \sim age + age^2 + age^3
$$

An example of the model fit is shown below:

```{r}
sumstats <- function(dat) {
    # split dat into 2 year age bins
    # dat <- dat %>% mutate(agebin = round(age / 2) * 2)
    dat$agebin <- round(dat$age)
    ss <- group_by(dat, agebin) %>%
        do({
            a <- summary(lm(y ~ g, data=.))
            tibble(
                agebin = .$agebin[1],
                beta = a$coefficients[2, 1],
                se = a$coefficients[2, 2]
            )
        })
    return(ss)

}

plot_gxage <- function(sim) {
    ss <- sumstats(sim$dat)
    m2 <- rma.uni(yi=beta, sei=se, mods = ~ poly(agebin, degree=3, raw=TRUE), data=ss)
    coefs <- tibble(
        method="metareg",
        poly=0:3,
        b = m2$beta[1:4],
        se = m2$se[1:4],
        pval = m2$pval[1:4]
    )
    sav <- predict(m2, newmods=unname(poly(ss$agebin, degree=3, raw=TRUE))) %>% as_tibble() %>% mutate(agebin=ss$agebin)
    plot(pred ~ agebin, data=sav, type="l")
    lines(ci.lb ~ agebin, data=sav, type="l", lty=2)
    lines(ci.ub ~ agebin, data=sav, type="l", lty=2)
    lines(bxage ~ age, sim$coef, col="red")
    points(beta ~ agebin, data=ss, pch=16, cex=0.01/ss$se)
    return(coefs)
}

plot_gxage(sim)
```

The red line is the true genotype x age coefficient, and the black line is the estimated effect of the genotype on the phenotype with age. Dotted lines are the confidence intervals.

In this example there is no heterogeneity in the effect of the genotype on the phenotype with age, so the confidence intervals of the polynomial fit are very wide.

Example **with interaction**:

```{r}
sim <- simdat(500000, 0, 0.02, 0.02, 0.02)
plot_gxage(sim)
```

Now the confidence intervals are much narrower, and the true effect of the genotype on the phenotype with age is captured by the polynomial fit.

## Performance simulations

Evaluate power and type 2 error of different approaches to detect G x age interactions. Approaches using age-stratified summary statistics:

- Cochran's Q statistic
- Meta-regression with simple linear model
- Meta-regression with polynomial terms

Approaches using individual level data across all ages:

- Main effect of genotype (no interaction)
- Linear G x age interaction
- Polynomial G x age interaction

```{r}
estimation <- function(dat) {
    ss <- sumstats(dat)
    m1 <- rma.uni(yi=ss$beta, sei=ss$se, method="FE")
    m2 <- rma.uni(yi=beta, sei=se, mods = ~ poly(agebin, degree=3), data=ss, method="FE")
    # m6 <- rma(yi=beta, sei=se, mods = ~ rcs(agebin, 4), data=ss)
    m7 <- rma(yi=beta, sei=se, mods = ~ agebin, data=ss, method="FE")

    # Compare against individual level models
    modpval <- function(mod) { pf(mod$fstat[1], mod$fstat[2], mod$fstat[3], lower.tail=FALSE) }

    reducedmod <- lm(y ~ g, data=dat)
    m5 <- anova(reducedmod, lm(y ~ g * age, data=dat))
    m3 <- anova(reducedmod, lm(y ~ g * poly(age, 3), data=dat))
    # m8 <- anova(reducedmod, lm(y ~ g * rcs(age, 4), data=dat))
    m9 <- anova(reducedmod, lm(y ~ g * as.factor(age), data=dat))

    bind_rows(
        tibble(method="main", pval=summary(reducedmod)$coefficients[2, 4]),
        tibble(method="ind_linear", pval=m5$P[2]),
        tibble(method="ind_poly", pval=m3$P[2]),
        # tibble(method="ind_rcs", pval=m8$P[2]),
        tibble(method="ind_full", pval=m9$P[2]),
        tibble(method="heterogeneity", Q=m1$QE, pval=m1$QEp),
        tibble(method="meta_linear", Q=m7$QM, pval=m7$QMp),
        tibble(method="meta_poly", Q=m2$QM, pval=m2$QMp),
        # tibble(method="meta_rcs", Q=m6$QM, pval=m6$QMp),
    )
}

# Function to run whole simulation
run_sim <- function(b0=0, b1=0, b2=0, b3=0, n=500000, sim=1) {
    args <- as.list(environment()) %>% as_tibble()
    sim <- simdat(n, b0, b1, b2, b3)
    est <- estimation(sim$dat)
    est <- bind_cols(est, as_tibble(args))
    return(est)
}

# Example
run_sim(0.02, 0, 0, 0.05)
dat <- simdat(500000, 0.02, 0, 0.02, 0.05)$dat

```

Do this over a range of scenarios (this was run on HPC in parallel)

```{r, eval=FALSE}
library(furrr)
param <- expand.grid(
    n = c(300000),
    b0 = c(0, -0.02),
    b1 = c(0, 0.02),
    b2 = c(0, -0.02, 0.02),
    b3 = c(0, -0.02, 0.02),
    sim=1:200
)

plan(multicore(workers=50))
opt <- furrr_options(seed=TRUE)
res <- future_pmap(param, run_sim, .progress=TRUE, .options=opt) %>% bind_rows()

saveRDS(res, "sim.rds")
```


## Results

```{r}
library(here)
res <- readRDS(here("posts/2024-11-23-age-interaction-power/sim.rds"))
```


**First check false discovery rates**

```{r}
res %>% group_by(main = b0 != 0, gxage = (b1 != 0 | b2 != 0 | b3 != 0), method) %>%
    summarise(power=mean(pval < 0.05)) %>%
    filter(!gxage) %>%
    ggplot(aes(x=method, y=power, fill=main)) +
    geom_bar(stat="identity", position="dodge") +
    geom_hline(yintercept=0.05, linetype="dashed") +
    labs(y="False discovery rate")
```

Model names:

- `main` = Individual level main effect of genotype with no interaction estimation
- `ind_linear` = Individual level G x age with linear interaction
- `ind_poly` = Individual level G x age with polynomial interaction
- `ind_full` = Individual level G x age with age as a categorical variable
- `heterogeneity` = Cochran's Q statistic for age-stratified summary statistics
- `meta_linear` = Meta-regression for age-stratified summary statistics using linear interaction
- `meta_poly` = Meta-regression for age-stratified summary statistics using polynomial interaction

Note that all models are wrt 

Summary:

- Heterogeneity has well controlled type 1 error when there is a main effect and when there is no main effect
- individual level and meta regression linear models both have well controlled type 1 error when there is no interaction
- individual level and meta regression polynomial models both have inflated type 1 error when there is no interaction (due to overfitting)

**Check power**

```{r}
library(ggplot2)
res %>% group_by(main = b0 != 0, gxage = (b1 != 0 | b2 != 0 | b3 != 0), method) %>%
    summarise(power=mean(pval < 0.05)) %>%
    filter(gxage) %>%
    ggplot(aes(x=method, y=power, fill=main)) +
    geom_bar(stat="identity", position="dodge")
```

Summary:

- Detecting G x age using `heterogeneity` (Cochran's Q) is reasonably powered but has lower power than other approaches
- Individual models have higher power than meta-regression models


---

```{r}
sessionInfo()
```
