---
title: "residual-height"
author: Gibran Hemani
date: "2026-02-13"
categories: []
---

## Background

$$
X = \beta_{gx} G + \beta_{cx} C + \epsilon_x
$$

$$
Y = \beta_{xy} X + \beta_{cy} C + \epsilon_y
$$

Two stage least squares aims to estimate the causal effect of X on Y, $\beta_{xy}$, by using G as an instrument for X. The first stage regression is:

$$
X = \beta_{gx} G
$$

Make a prediction of X from G, and then use this predicted value of X in the second stage regression to estimate $\beta_{xy}$.  

$$
\hat{X} = \hat{\beta}_{gx} G
$$

And then the second stage regression is:

$$
Y = \beta_{xy} \hat{X} + error
$$

So $\hat{\beta_{xy}}$ is the estimate of the causal effect of X on Y, which is independent of the common confounding of X and Y.

## Simulation

Generate Height:

```{r}
set.seed(123)
library(dplyr)
library(ggplot2)
library(furrr)
library(knitr)
n <- 1000000
h2 <- 0.7
r2_prs <- 0.75
prs_explained <- rnorm(n, mean = 0, sd = sqrt(r2_prs * h2))
prs_unexplained <- rnorm(n, mean = 0, sd = sqrt((1 - r2_prs) * h2))
childhood_environment <- rnorm(n, mean = 0, sd = sqrt(1 - h2))
height <- prs_explained + prs_unexplained + childhood_environment
residual_height <- height - prs_explained
```

Generate coronary heart disease (just a continuous measure for simplicity):

```{r}
beta_height_chd <- -0.1
beta_environment_chd <- -0.2
chd <- beta_height_chd * height + beta_environment_chd * childhood_environment + rnorm(n)
```

Get estimates of the effect of PRS, height and residual height on CHD:

```{r}
get_est <- function(x, y) {
    mod <- summary(lm(y ~ x))$coefficients
    tibble(
        beta = mod[2, 1],
        se = mod[2, 2],
        p = mod[2, 4],
        rsq = cor(x, y)^2,
        var_exposure = var(x)
    )
}
bind_rows(
    get_est(prs_explained, chd) %>% mutate(exposure = "PRS"),
    get_est(height, chd) %>% mutate(exposure = "Height"),
    get_est(residual_height, chd) %>% mutate(exposure = "Residual Height")
) %>% kable()
```

Generalise this simulation to a function:

```{r}
sim <- function(h2, r2_prs, beta_height_chd, beta_environment_chd, n = 100000, simrep = 1) {
    args <- list(
        h2 = h2,
        r2_prs = r2_prs,
        beta_height_chd = beta_height_chd,
        beta_environment_chd = beta_environment_chd,
        n = n,
        simrep = simrep
    )
    prs_explained <- rnorm(n, mean = 0, sd = sqrt(r2_prs * h2))
    prs_unexplained <- rnorm(n, mean = 0, sd = sqrt((1 - r2_prs) * h2))
    childhood_environment <- rnorm(n, mean = 0, sd = sqrt(1 - h2))
    height <- prs_explained + prs_unexplained + childhood_environment
    residual_height <- height - prs_explained
    chd <- beta_height_chd * height + beta_environment_chd * childhood_environment + rnorm(n)
    res <- bind_rows(
        get_est(prs_explained, chd) %>% mutate(exposure = "PRS"),
        get_est(height, chd) %>% mutate(exposure = "Height"),
        get_est(residual_height, chd) %>% mutate(exposure = "Residual Height")
    )
    bind_cols(args, res)
}
sim(h2 = 0.7, r2_prs = 0.75, beta_height_chd = -0.1, beta_environment_chd = -0.1) %>% kable()
```

Try a few different scenarios

```{r}
#| warning: false
params <- expand.grid(
    h2 = 0.7,
    r2_prs = c(0.75, 0.4),
    beta_height_chd = c(0, -0.1, -0.2, -0.3),
    beta_environment_chd = c(-0.1, -0.2, -0.3),
    simrep = 1:10
)

plan(multicore)
results <- future_pmap(params, sim, .progress = TRUE) %>% bind_rows() 
```

look at the results:

```{r}
ggplot(results, aes(x = beta_height_chd, y = beta, colour = exposure)) +
    geom_point() +
    facet_grid(r2_prs ~ beta_environment_chd, labeller = label_both) +
    scale_colour_brewer(type="qual")
```


## Estimate effect of environment on Y

The height residual is still heritable. We can estimate the path from E -> X -> Y because

- E->X can be obtained with MR of residual -> height
- X->Y can be obtained with MR of height -> CHD

```{r}
est_ey <- function(dat) {
    
    dat <- dat[[1]]
    xhat <- fitted.values(lm(height ~ prs_explained, dat))
    tibble(
        cov_prs_chd = cov(dat$prs_explained, dat$chd),
        cov_xhat_chd = cov(xhat, dat$chd),
        cov_x_chd = cov(dat$height, dat$chd),
        var_x = var(dat$height),
        var_prs = var(dat$prs_explained),
        var_chd = var(dat$chd),
        var_environment = var(dat$childhood_environment),
        var_xhat = var(xhat),
        lm_x_chd = lm(chd ~ height, dat)$coef[2],
        lm_xhat_chd = lm(dat$chd ~ xhat)$coef[2]
    )
}
```


## Can we estimate the effect of environment on Y using the residual height?

According to the model described above, the influence of adverse childhood environment on CHD is due to two paths:

- $\beta_{cy}$: the direct effect of environment on CHD
- $\beta_{cx} \beta_{xy}$: the effect of environment on CHD mediated by height

The PRS effect estimate on CHD represents $\beta_{xy}$, the causal effect of height on CHD. Can we express $\beta_{cy}$ in terms of values that are observable in the data?

$$
\begin{aligned}
cov(X, Y) &= cov(\beta_{gx} G + \beta_{cx} C + \epsilon_x, \beta_{xy} X + \beta_{cy} C + \epsilon_y) \\
&= \beta_{gx} \beta_{xy} var(G) + (\beta_{cx} \beta_{xy} + \beta_{cy}) var(C) + \beta_{xy} var(\epsilon_x) \\
&= \beta_{xy} var(X) h^2 + (\beta_{cx}^2 \beta_{xy} + \beta_{cx} \beta_{cy}) var(C) + \beta_{xy} var(X)((1-h^2) - var(C)) \\
\end{aligned}
$$

This can be rearranged to be find an expression for $\beta_{cy}$:

$$
\begin{aligned}
\beta_{cy} &= \frac{cov(X, Y) - \beta_{xy} var(X) h^2 - [(1-h^2) var(X) - \beta_{cx}^2 var(C)] \beta_{xy} - \beta_{cx}^2 \beta_{xy} var(C)}{var(C) \beta_{cx}} \\
&= \frac{cov(X, Y) = \beta_{xy} var(X)}{var(C) \beta_{cx}}
\end{aligned}
$$

In principle we can also express $\beta_{cy}$ in terms of the covariance of residual height association with CHD, which is observable in the data:

$$
\beta_{cy} = \frac{cov(X, Y | \hat{G}) - \beta_{xy} (var(X)(1-h^2) + h^2 var(X)(1-R^2))}{var(C) \beta_{cx}}
$$

where $R^2$ is the proportion of the heritability of height explained by the PRS.

So we should be able to check empirically that 

$$
cov(X, Y | \hat{G}) - \beta_{xy} (var(X)(1-h^2) + h^2 var(X)(1-R^2)) = cov(X, Y) = \beta_{xy} var(X)
$$

In order to estimate $\beta_{cy}$, we need to know two parameters:

- $var(C)$: the variance of 'childhood environment'. It is essentially $var(C) = (1 - h^2) var(X) - var(\epsilon_x)$, where $var(\epsilon_x)$ is the developmental stochastic process, leaving $var(C)$ to represent the shared and non-shared environment. Empirically, this is estimated at around 10% of the variance in height, so we can set $var(C) = 0.1 var(X)$.
- $\beta_{cx}$: the effect of childhood environment on height. This cannot be easily estimated, so we can set this to 1, which means that when we interpret $\beta_{cy}$, we are interpreting it as the effect of childhood environment on CHD relative to its effect on height.

## Function to estimate the effect of environment on Y

```{r}
est_cy <- function(height, prs, outcome, h2 = 0.7, prop_height_env = 0.1) {
    var_x <- var(height)
    var_prs <- var(prs)
    var_c <- prop_height_env * var_x
    cov_x_y <- cov(height, outcome)

    return(beta_cy)
}
```


```{r}
library(AER) # For ivreg

est_cy <- function(height, prs, outcome, h2 = 0.7, prop_height_env = 0.1, n_boot = 1000) {
  
  r2_prs <- cor(prs, height)^2

  # Filter out NAs for a consistent sample size
  df <- tibble(height, prs, outcome)
  df <- na.omit(df)
  n_obs <- nrow(df)
  
  # First-Stage F-statistic
  first_stage <- lm(height ~ prs, data = df)
  f_stat <- summary(first_stage)$fstatistic[1]
  
  # Two-Stage Least Squares (2SLS)
  tsls_model <- AER::ivreg(outcome ~ height | prs, data = df)
  b_xy <- coef(tsls_model)["height"]
  se_b_xy <- summary(tsls_model)$coefficients["height", "Std. Error"]
  
  calc_buy_total <- function(c_xy, v_x, b_val, vu) {
    (c_xy - (b_val * v_x)) / (1 * vu)
  }
  
  calc_buy_res <- function(c_res_xy, v_x, b_val, h_sq, v_u, r2_prs) {
    (c_res_xy - b_val * (v_x * (1 - h_sq) + h_sq * v_x * (1 - r2_prs))) / (v_u * 1)
  }
  
  # Point Estimates
  var_x <- var(df$height)
  var_u <- prop_height_env * var_x
  cov_x_y <- cov(df$height, df$outcome)
  
  res_x <- residuals(first_stage)
  cov_res_x_y <- cov(res_x, df$outcome)
  
  b_uy_total <- calc_buy_total(cov_x_y, var_x, b_xy, var_u)
  b_uy_res   <- calc_buy_res(cov_res_x_y, var_x, b_xy, h2, var_u, r2_prs)
  
  # Bootstrapping for SEs and CIs
  boot_results <- replicate(n_boot, {
    idx <- sample(1:n_obs, replace = TRUE)
    df_b <- df[idx, ]
    
    # Bootstrap 2SLS
    m_b <- AER::ivreg(outcome ~ height | prs, data = df_b)
    b_xy_b <- coef(m_b)["height"]
    
    var_x_b <- var(df_b$height)
    var_u_b <- prop_height_env * var_x_b
    
    # Total Cov method
    c_xy_b <- cov(df_b$height, df_b$outcome)
    buy_t <- calc_buy_total(c_xy_b, var_x_b, b_xy_b, var_u_b)
    
    # Residual Cov method
    m_res_b <- lm(height ~ prs, data = df_b)
    c_res_b <- cov(residuals(m_res_b), df_b$outcome)
    buy_r <- calc_buy_res(c_res_b, var_x_b, b_xy_b, h2, var_u_b, r2_prs)
    
    c(b_xy_b, buy_t, buy_r)
  })
  
  # Calculate Stats
  estimates <- c(b_xy, b_uy_total, b_uy_res)
  # b_xy SE is from AER package, b_uy SEs from bootstrap
  ses <- c(se_b_xy, sd(boot_results[2,]), sd(boot_results[3,]))
  
  # P-values and CIs
  z_scores <- estimates / ses
  p_vals <- 2 * (1 - pnorm(abs(z_scores)))
  
  ci_lower <- c(estimates[1] - 1.96 * ses[1], 
                quantile(boot_results[2,], 0.025), 
                quantile(boot_results[3,], 0.025))
  
  ci_upper <- c(estimates[1] + 1.96 * ses[1], 
                quantile(boot_results[2,], 0.975), 
                quantile(boot_results[3,], 0.975))
  
  # Final Table
  results_table <- tibble(
    Parameter = c("b_xy (2SLS)", "b_uy (Total Cov)", "b_uy (Resid Cov)"),
    Estimate = estimates,
    Std_Error = ses,
    P_value = p_vals,
    LCI_95 = ci_lower,
    UCI_95 = ci_upper,
    Sample_Size = n_obs,
    First_Stage_F = c(f_stat, NA, NA)
  )
  
  return(results_table)
}
```


h2 = var_prs * b^2 / var_x
r2env_height = var_u * b_ux^2 / var_x

var_x = var_prs * b^2 / h2
var_x = var_u * b_ux^2 / r2env_height



```{r}
sim2 <- function(b_xy, b_uy, b_ux=1, b_prs=1, h2=0.7, r2_env_height=0.1, var_u = 1, var_x=1, var_ey = 1, r2_prs=1, n = 100000) {

    r2_xe <- 1 - h2 - r2_env_height
    var_prs <- h2 * var_x / b_prs^2
    var_u <- r2_env_height * var_x / b_ux^2
    var_ex <- r2_xe * var_x

    args <- list(
        h2 = h2,
        r2_prs = r2_prs,
        b_xy = b_xy,
        b_uy = b_uy,
        b_ux = b_ux,
        b_prs = b_prs,
        v_ex = var_ex,
        v_ey = var_ey,
        v_u = var_u,
        v_prs = var_prs,
        n = n
    )
    dat <- tibble(
        prs_explained = rnorm(n, mean = 0, sd = sqrt(args$v_prs * b_prs * r2_prs)),
        prs_unexplained = rnorm(n, mean = 0, sd = sqrt(args$v_prs * b_prs * (1 - r2_prs))),
        prs = prs_explained + prs_unexplained,
        u = rnorm(n, mean = 0, sd = sqrt(args$v_u)),
        ex = rnorm(n, mean = 0, sd = sqrt(args$v_ex)),
        x = prs + u * b_ux + ex,
        residual_x = x - prs_explained,
        ey = rnorm(n, mean = 0, sd = sqrt(args$v_ey)),
        y = b_xy * x + b_uy * u + ey
    )
    return(list(args, dat))
}

dat <- sim2(b_xy = -0.2, b_uy = -0.2, r2_prs = 0.4, h2=0.7, n=1000000)[[2]]
est_cy(dat$x, dat$prs_explained, dat$y, h2 = 0.7, prop_height_env = 0.1, n_boot = 10)
```


## Simulations to check equations

Ignore this section for now, it's just me checking the equations are correct.

```{r}
library(dplyr)
sim2 <- function(b_xy, b_uy, b_ux=1, b_prs=1, h2=0.7, r2_env_height=0.1, var_u = 1, var_x=1, var_ey = 1, r2_prs=1, n = 100000) {
    r2_xe <- 1 - h2 - r2_env_height
    var_prs <- h2 * var_x / b_prs^2
    var_u <- r2_env_height * var_x / b_ux^2
    var_ex <- r2_xe * var_x

    args <- list(
        h2 = h2,
        r2_prs = r2_prs,
        b_xy = b_xy,
        b_uy = b_uy,
        b_ux = b_ux,
        b_prs = b_prs,
        v_ex = var_ex,
        v_ey = var_ey,
        v_u = var_u,
        v_prs = var_prs,
        n = n
    )
    dat <- tibble(
        prs_explained = rnorm(n, mean = 0, sd = sqrt(args$v_prs * b_prs * r2_prs)),
        prs_unexplained = rnorm(n, mean = 0, sd = sqrt(args$v_prs * b_prs * (1 - r2_prs))),
        prs = prs_explained + prs_unexplained,
        u = rnorm(n, mean = 0, sd = sqrt(args$v_u)),
        ex = rnorm(n, mean = 0, sd = sqrt(args$v_ex)),
        x = prs + u * b_ux + ex,
        residual_x = x - prs_explained,
        ey = rnorm(n, mean = 0, sd = sqrt(args$v_ey)),
        y = b_xy * x + b_uy * u + ey
    )
    return(list(args, dat))
}

est_obs <- function(simdat) {
    dat <- simdat[[2]]
    xhat <- fitted.values(lm(x ~ prs_explained, dat))
    residual <- residuals(lm(x ~ prs_explained, dat))
    h2 <- simdat[[1]]$h2
    r2_prs <- simdat[[1]]$r2_prs
    tibble(
        var_x = var(dat$x),
        var_prs_explained = var(dat$prs_explained),
        var_prs_unexplained = var(dat$prs_unexplained),
        var_prs = var(dat$prs_explained + dat$prs_unexplained),
        var_residual = var(dat$residual_x),
        var_y = var(dat$y),
        var_u = var(dat$u),
        var_xhat = var(xhat),
        var_ex = var(dat$ex),

        cov_prs_y = cov(dat$prs_explained, dat$y),
        cov_prs_x = cov(dat$prs_explained, dat$x),
        cov_xhat_y = cov(xhat, dat$y),
        cov_residual_y = cov(residual, dat$y),
        cov_x_y = cov(dat$x, dat$y),

        lm_x_y = lm(y ~ x, dat)$coef[2],
        lm_xhat_y = lm(dat$y ~ xhat)$coef[2],
        lm_residual_y = lm(dat$y ~ residual)$coef[2],
        lm_u_y = simdat[[1]]$b_uy,
        lm_u_y_resx = simdat[[1]]$b_uy,
        est_u_y = (cov_x_y - cov_prs_y) / (var_x - var_prs),
        numerator = (cov_x_y - lm_xhat_y * var_x),
        denominator = (cov_residual_y - lm_xhat_y * (var_x * (1 - h2) + h2 * var_x * (1 - r2_prs))),
        ratio = numerator / denominator
    ) %>% bind_cols(simdat[[1]]) %>% as.list()
}

est_exp <- function(res_obs) {

    a <- res_obs %>% as_tibble() %>% mutate(
        exp_var_x = var_prs * b_prs^2 + var_u * b_ux^2 + v_ex,
        exp_var_residual = exp_var_x - var_prs * b_prs^2 + var_prs * b_prs^2 * (1-r2_prs),
        exp_var_prs = h2 * exp_var_x,
        exp_var_prs_explained = h2 * r2_prs * exp_var_x,
        exp_var_prs_unexplained = h2 * (1 - r2_prs) * exp_var_x,
        exp_var_residual = exp_var_x - exp_var_prs_explained,
        exp_var_ex = exp_var_x * (1 - h2) - var_u * b_ux^2,
        # exp_var_y = b_xy^2 * exp_var_prs + (b_ux * b_xy + b_uy)^2 * var_u + v_ey + b_xy^2 * exp_var_ex,
        exp_var_y = b_xy^2 * h2 * exp_var_x + (b_ux * b_xy + b_uy)^2 * var_u + v_ey + b_xy^2 * (exp_var_x * (1-h2) - v_u * b_ux^2),

        # exp_cov_x_y = var_prs * b_prs^2 * b_xy + var_u * (b_ux^2 * b_xy + b_ux * b_uy) + v_ex * b_xy,
        exp_cov_x_y = b_xy * h2 * exp_var_x + (b_ux^2 * b_xy + b_uy * b_ux) * v_u + v_ex * b_xy,
        exp_cov_x_y = b_xy * h2 * exp_var_x + (b_ux^2 * b_xy + b_uy * b_ux) * v_u + (exp_var_x * (1-h2) - v_u * b_ux^2) * b_xy,
        exp_cov_prs_y = exp_var_prs_explained * b_prs * b_xy,
        exp_cov_prs_x = exp_var_prs_explained * b_prs,
        # exp_cov_residual_y = var_u * (b_ux^2 * b_xy + b_ux * b_uy) + v_ex * b_xy + (var_prs * b_prs * b_xy) * (1-r2_prs),
        exp_cov_residual_y = var_u * (b_ux^2 * b_xy + b_ux * b_uy) + (exp_var_x * (1-h2) - v_u * b_ux^2) * b_xy + (h2 * exp_var_x * b_prs * b_xy) * (1-r2_prs),

        exp_lm_x_y = exp_cov_x_y / exp_var_x,        
        exp_lm_residual_y = exp_cov_residual_y / exp_var_residual,
        exp_lm_xhat_y = exp_cov_prs_y / exp_cov_prs_x,
        # exp_lm_u_y = (exp_cov_x_y - (b_xy * h2 * exp_var_x) - (exp_var_x * (1 - h2) - v_u * b_ux^2) * b_xy - (b_ux^2 * b_xy) * v_u) / (b_ux * v_u)
        exp_lm_u_y = (exp_cov_x_y - (b_xy * exp_var_x)) / (b_ux * v_u),
        exp_lm_u_y_resx = (exp_cov_residual_y - b_xy * (exp_var_x * (1 - h2) + h2 * exp_var_x * b_prs * (1 - r2_prs))) / (v_u * b_ux),
        exp_ratio = exp_lm_u_y / exp_lm_u_y_resx
    )
    b <- a %>% select(starts_with("exp"))
    names(b) <- gsub("exp_", "", names(b))
    b <- tibble(parameter = names(b), exp = as.numeric(b[1, ]))

    a <- a %>% select(-starts_with("exp"))
    a <- tibble(parameter = names(a), obs = as.numeric(a[1, ]))

    inner_join(a, b, by = "parameter") %>% mutate(ratio = round(obs / exp, 2)) %>% as.data.frame
}

res <- sim2(-0.5, -0.2, 1, r2_prs = 0.7, h2=0.7, n=1000000) %>% est_obs %>% est_exp
res
```



---

```{r}
sessionInfo()
```
