---
title: "Type 2 diabetes risk differences between populations"
author: Gibran Hemani
date: "2023-11-21"
categories: []
---

## Background

T2D rates are ~5x higher amongst South Asians than native British populations in Britain. Most likely due to environmental differences, but could it hypothetically be explained by differences in genetic architecture?

Assume polygenic common variant distribution and a disease liability threshold model.

## Descriptive statistice

- t2d diagnosis normally 45-65
- people aged 45+ constitute approx 50% of 70 million people
- therefore 4.9 million t2d cases in uk out of 35 million people who are susceptible
- 3.07 million South Asians, assume 50% aged 45+
- Prevalence is 5x higher amongst South Asians

Simultaneous equations:

$$
4.9 = 1.54 x + 35y
$$

where $x$ is the prevalence amongst South Asians and $y$ is the prevalence amongst white British

Also

$$
x = y * 5
$$

Calculate prevalences:

```{r}
A <- matrix(c(1.54, 35, -1, 5), nrow=2, byrow=T)
b <- c(4.9, 0)
prev <- solve(A, b)
prev
```

## Model

Assume threshold is the same in both populations. Establish a threshold based on European genetic architecture

```{r}
library(ggplot2)
library(dplyr)

sim1 <- function(mg, vg, me, ve, prev) {
    n <- 100000
    g <- rnorm(n, mg, sqrt(vg))
    e <- rnorm(n, me, sqrt(ve))
    l <- g + e
    thresh <- quantile(l, 1-prev)
    return(thresh)
}

thresh <- sim1(0, 0.4, 0, 0.6, prev[2])
thresh
```

Estimate prevalence given a threshold when parameters change

```{r}

sim2 <- function(mg, vg, me, ve, thresh) {
    n <- 100000
    g <- rnorm(n, mg, sqrt(vg))
    e <- rnorm(n, me, sqrt(ve))
    l <- g + e
    prev <- sum(l > thresh) / n
    return(prev)
}

sim2(0.1, 0.4, 0, 0.6, thresh)
```

Enumerate possibilities

```{r}
param <- expand.grid(
    mg = seq(0, 4, by=0.01),
    vg = 0.4,
    me = 0,
    ve = 0.6,
    thresh = thresh,
    prev = NA
)

for(i in 1:nrow(param)) {
    param$prev[i] <- sim2(param$mg[i], param$vg[i], param$me[i], param$ve[i], param$thresh[i])
}

param$ratio = param$prev / prev[2]

ggplot(param, aes(x=mg, y=ratio)) +
geom_line() +
geom_hline(yintercept=5, linetype="dotted")
```

This shows how much larger the mean of the genetic liability would have to be to lead to 5x prevalence

Comparison of distributions:

```{r}
mg <- param$mg[which(param$ratio >= 5)[1]]
bind_rows(
    tibble(pop="EUR", x=seq(-3, 3, length.out=1000), val = dnorm(x, sd=sqrt(0.4))),
    tibble(pop="SAS", x=seq(-3+mg, 3+mg, length.out=1000), val = dnorm(x, m=mg, sd=sqrt(0.4)))
) %>% ggplot(., aes(x=x, y=val)) +
geom_line(aes(colour=pop)) +
geom_vline(xintercept=thresh)
```


Is it plausible in terms of just different allele frequencies?

```{r}
nsnp <- 10000
af <- rbeta(nsnp, 1, 1) / 2
hist(af)
b <- rnorm(nsnp, 0, (2*af*(1-af))^-0.4)
vg <- sum(b^2 * 2 * af * (1-af))
b <- b / sqrt(vg) * 0.4
plot(b ~ af)

nid <- 1000
g1 <- sapply(af, \(x) { rbinom(nid, 2, x) })
score1 <- g1 %*% b
hist(score1)
sd(score1)

```


## Summary

- Seems unlikely under this model

---

```{r}
sessionInfo()
```
