---
title: "Heterogeneity in cell-type interaction model"
author: Gibran Hemani
date: "2023-11-28"
categories: []
---

## Background

Two models

- Estimate mQTL in purified cells. Marginal effects obtained per cell type. Can estimate heterogeneity between of mQTL effect estimates across cell types
- Estimate G x celltype proportion interaction for each celltype. Interaction is a deviation fron mean effect. Quite a complex model Can we just estimate heterogeneity between interaction terms?

```{r}
library(ggplot2)
library(dplyr)

sim <- function(nc, n, betas = runif(nc, -2, 2)) {
    g <- rbinom(n, 2, 0.4)
    
    m <- sapply(1:nc, function(i)
    {
        g * betas[i] + rnorm(n)
    })
    # for each individual sample cell type proportions
    cellprop <- sapply(1:n, function(x) {a <- runif(nc); a/sum(a)}) %>% t()
    # weighted sum
    M <- (scale(m) * cellprop) %>% rowSums

    res <- lapply(1:nc, function(i)
    {
        o <- summary(lm(M ~ g * cellprop[,i]))
        tibble(cell=i, b=o$coef[4,1], se=o$coef[4,2], pval=o$coef[4,4], method="interaction")
    }) %>% bind_rows()

    res2 <- lapply(1:nc, function(i)
    {
        o <- summary(lm(m[,i] ~ g))
        tibble(cell=i, b=o$coef[2,1], se=o$coef[2,2], pval=o$coef[2,4], method="marginal")
    }) %>% bind_rows()

    betas <- tibble(cell=1:nc, b=betas, method="true")
    

    # cellprop <- as_tibble(cellprop)
    # m <- as_tibble(m)
    # names(m) <- paste0("m", 1:nc)
    # names(cellprop) <- paste0("cp", 1:nc)
    # cellprop$M <- M
    # cellprop$g <- g

    # cellprop <- bind_cols(cellprop, m)

    # return(tibble(int=res, marg=res2, betas))


    # heterogeneity
    het <- bind_rows(
        fixed_effects_meta_analysis(res$b, res$se) %>% 
            as_tibble() %>%
            slice_head(n=1) %>%
            mutate(method="interaction"),
        fixed_effects_meta_analysis(res2$b, res2$se) %>% 
            as_tibble() %>%
            slice_head(n=1) %>%
            mutate(method="marginal"),
    )

    effs <- bind_rows(res, res2, betas)
    return(list(effs=effs, het=het))
}
```

```{r}
# Provide vector of interaction betas and their standard errors
fixed_effects_meta_analysis <- function(beta_vec, se_vec) {
    w <- 1 / se_vec^2
    beta <- sum(beta_vec * w, na.rm=T) / sum(w, na.rm=T)
    se <- sqrt(1 / sum(w, na.rm=T))
    pval <- pnorm(abs(beta / se), lower.tail = FALSE)
    Qj <- w * (beta-beta_vec)^2
    Q <- sum(Qj, na.rm=T)
    Qdf <- sum(!is.na(beta_vec))-1
    if(Qdf == 0) Q <- 0
    Qjpval <- pchisq(Qj, 1, lower.tail=FALSE)
    Qpval <- pchisq(Q, Qdf, lower.tail=FALSE)
    pv <- pnorm(abs(beta_vec)/se_vec, lower.tail=FALSE)
    min_pv <- min(pv, na.rm=T)
    return(list(beta=beta, se=se, Q=Q, Qdf=Qdf, Qpval=Qpval, Qj=Qj, Qjpval=Qjpval, min_pv=min_pv))
}
```

```{r}
param <- expand.grid(
    nc=5,
    n=seq(100,1000, by=50),
    bsd=0.4,
    nsim=1:100
)
dim(param)
r <- lapply(1:nrow(param), function(i) {
    o <- sim(param$nc[i], param$n[i], rnorm(param$nc[i], sd=param$bsd[i]))
    bind_rows(
        param[i,] %>% mutate(Qpval=o$het$Qpval[1], method="interaction"),
        param[i,] %>% mutate(Qpval=o$het$Qpval[2], method="marginal")
    )
}) %>% bind_rows()

r %>% group_by(n, method) %>%
    summarise(pow=sum(Qpval < 0.05)/n()) %>%
    ggplot(., aes(x=n, y=pow)) +
    geom_point(aes(colour=method)) +
    geom_line(aes(colour=method))
```

```{r}
ggplot(r, aes(x=as.factor(n), y=-log10(Qpval))) +
    geom_boxplot(aes(colour=method))
```

## Summary

- Heterogeneity estimate seems ok
- Interaction model is very low power compared to marginal cell estimates

---

## Ignore

```{r, eval=FALSE}
d <- sim(2, 1000)

sim(5, 1000, b=rnorm(5, sd=0.4))
sim(2, 10000, c(1,2))


n <- 10000
nc <- 2
betas <- c(1,2)
g <- rbinom(n, 2, 0.4)

m <- sapply(1:nc, function(i) {
    g * betas[i] + rnorm(n)
})
# for each individual sample cell type proportions
cp <- sapply(1:n, function(x) {a <- runif(nc); a/sum(a)}) %>% t()
# weighted sum
M <- (scale(m) * cp) %>% rowSums

cov(M, g * cp[,1]) / var(g*cp[,1])

summary(lm(M ~ g*cp[,1]))

summary(lm(M ~ g))
cov(M, g) / var(g)

summary(lm(M ~ cp[,1]))
cov(M, cp[,1]) / var(cp[,1])

summary(lm(M ~ g*cp[,1]))
cov(M, g*cp[,1]) / var(g*cp[,1])

Mr <- residuals(lm(M ~ g+cp[,1]))
summary(lm(Mr ~ g:cp[,1]))
cov(Mr, g*cp[,1]) / var(g*cp[,1])




summary(lm(M ~ g:cp[,1]))
cov(M, cp[,1]) / var(cp[,1])


summary(lm(M ~ g * cp1, d))
summary(lm(m1 ~ g, d))
```

```{r}
sim2 <- function(nc, n, betas = runif(nc, -2, 2)) {
    g <- rbinom(n, 2, 0.4)
    
    m <- sapply(1:nc, function(i)
    {
        g * betas[i] + rnorm(n)
    })
    # for each individual sample cell type proportions
    cellprop <- sapply(1:n, function(x) {a <- runif(nc); a/sum(a)}) %>% t()
    # weighted sum
    M <- (scale(m) * cellprop) %>% rowSums

    res <- lapply(1:nc, function(i)
    {
        o <- summary(lm(M ~ g * cellprop[,i]))
        tibble(cell=i, b=o$coef[4,1], se=o$coef[4,2], pval=o$coef[4,4], method="interaction")
    }) %>% bind_rows()

    res2 <- lapply(1:nc, function(i)
    {
        o <- summary(lm(m[,i] ~ g))
        tibble(cell=i, b=o$coef[2,1], se=o$coef[2,2], pval=o$coef[2,4], method="marginal")
    }) %>% bind_rows()

    betas <- tibble(cell=1:nc, b=betas, method="true")
    

    cellprop <- as_tibble(cellprop)
    m <- as_tibble(m)
    names(m) <- paste0("m", 1:nc)
    names(cellprop) <- paste0("cp", 1:nc)
    cellprop$M <- M
    cellprop$g <- g

    cellprop <- bind_cols(cellprop, m)
    return(cellprop)
}

ana <- function(d) {
    nc <- length(grep("cp", names(d)))

    # per cell estimates
    o1 <- lapply(1:nc, \(x) { 
        f <- paste0("M ~ g * cp", x)
        o <- summary(lm(f, data=d))$coef
        n <- rownames(o)
        o <- as_tibble(o)
        names(o) <- c("b", "se", "tval", "pval")
        o$term <- n
        o$cell <- x
        o$mod <- "interaction"
        o
    }) %>% bind_rows()

    # cell int estimates
    o2 <- lapply(1:nc, \(x) { 
        f <- paste0("m", x, " ~ g")
        o <- summary(lm(f, data=d))$coef
        n <- rownames(o)
        o <- as_tibble(o)
        names(o) <- c("b", "se", "tval", "pval")
        o$term <- n
        o$cell <- x
        o$mod <- "marginal"
        o
    }) %>% bind_rows()
    bind_rows(o1, o2)
}
d <- sim2(2, 1000)
ana(d) %>% as.data.frame
```

```{r, eval=FALSE}
cov(residuals(lm(M ~ g + cp1, d)), d$g*d$cp1) / var(d$g*d$cp1)
cov(d$M, d$g*d$cp1) / var(d$g*d$cp1)
summary(lm(M ~ g*cp1, d))
cov(d$M, d$g) / var(d$g)

summary(lm(M ~ g + cp1, d))
summary(lm(M ~ g + cp1 + g*cp1, d))
summary(lm(M ~ g + cp1 + g*cp1 + cp2 + g*cp2, d))


r <- residuals(lm(M ~ g + cp1, d))

summary(lm(r ~ g*cp1, d))
cov(d$cp1)

n <- 10000
a <- rnorm(n, m=0 sd=10)
b <- rnorm(n, sd=3)
y <- a + b - 4*a*b + rnorm(n)

cov(y, a*b) / var(a*b)
summary(lm(y ~ a*b))
```

---

```{r}
sessionInfo()
```
