---
title: "Rsq in simulations"
author: Gibran Hemani
date: "2023-12-06"
categories: []
---

## Background

What determines $R^2$ between X and Y in a linear model under confounding

$$
\begin{aligned}

Y &= a + bX + E \\
b &= cov(X, Y) / var(X) \\
R &= cov(X, Y) / [sd(X)sd(Y)] \\
&= b * sd(x) / sd(y)

\end{aligned}
$$

In OLS the b will be different to the IV b if there is unmeasured confounding

$$
\begin{aligned}
X &= a + \beta_1 G + \beta_2 U + \epsilon \\
Y &= a + b_1 X + b_2 U + E \\

b_{OLS} &= cov(X, Y) / var(X) \\
&= cov(\beta_1 G + \beta_2 U, b_1(\beta_1 G + \beta_2 U) + b_2 U)/var(X) \\
&= cov(\beta_1 G + \beta_2 U, b_1\beta_1 G + b_1\beta_2 U + b_2 U)/var(X) \\
&= [b_1\beta_1^2 var(G) + (b_1\beta_2^2 + b_2\beta_2) var(U)]/var(X) \\


\end{aligned}
$$

and

$$
R_{OLS} = b_{OLS} sd(x)/sd(y)
$$

therefore to get the OLS $R^2$ of X,Y

$$
R^2 = \left [ \frac{b_1\beta_1^2 var(G) + (b_1\beta_2^2 + b_2\beta_2) var(U)}{sd(x) sd(y)} \right]^2
$$

Note that 

$$
\begin{aligned}
var(x) &= sd(x)^2 = \beta_1^2var(G) + \beta_2^2var(U) + var(\epsilon) \\
var(y) &= sd(y)^2 = b_1^2var(X) + b_2^2var(U) + var(E) \\
\end{aligned}
$$

So ultimately if you want to fix $R^2$ for different parameters of effects you should be able to scale $var(\epsilon)$ and $var(E)$, the residual variances, according to these formulae. 

By contrast the variance explained by the causal effect of X is

$$
R^2_{IV, x,y} = b^2_1var(X) / var(Y)
$$


## Check

```{r}
set.seed(1)
b1 <- 0.2
b2 <- 3
beta1 <- 4
beta2 <- 5
n <- 10000
u <- rnorm(n)
g <- rnorm(n)
x <- u * beta2 + g * beta1 + rnorm(n)
y <- u * b2 + x * b1 + rnorm(n, sd=0)
```

Beta

```{r}
cov(x, y)/var(x)
summary(lm(y ~ x))$coef[2,1]
( b1*beta1^2*var(g) + (b1*beta2^2 + b2*beta2) * var(u) ) / var(x)
```

Correlation

```{r}
((( b1*beta1^2*var(g) + (b1*beta2^2 + b2*beta2) * var(u) ) / var(x)) * sd(x) / sd(y))^2
cor(x, y)^2
```

---

```{r}
sessionInfo()
```
