---
title: "IV confounding with changing variances"
author: Gibran Hemani
date: "2023-07-20"
categories: []
---

## Background

When instruments arise through U the bias is buy/bux, whereas ols bias is buy*bux. If buy is larger than bux then the iv bias will be smaller than ols bias, but otherwise likely to be larger than ols bias.

What happens if you just rescale the values of x and y?


bias_ols = buy * bux

When the SNP goes through U
bias_iv = buy/bux


buy = 0.1
bux = 0.1

bias_iv = 1
bias_ols = 0.01

---

change sd of y and x to be from 1 to 10

buy = 1
bux = 1

bias_iv = 1
bias_ols = 1

---

buy = 0.1
bux = 1

bias_iv = 0.1
bias_ols = 0.1


y = buy * u + e
x = bux * u + e

b_ols = cov(buy*u, bux*u)
= buy * bux * var(u)


```{r}
library(ggplot2)
library(tidyr)
library(dplyr)
library(TwoSampleMR)
library(simulateGP)

iv_bias <- function(g, x, y) {
    bgx = cov(g, x) / var(g)
    bgy = cov(g, y) / var(g)
    return(bgy/bgx)
}

ols_bias <- function(x, y) {
    return(cov(x,y)/var(x))
}

n <- 100000
g <- rbinom(n, 2, 0.4)
u <- g + rnorm(n, 0, sqrt(1 - var(g)))

bux <- 0.5
buy <- 0.5
y <- u * buy + rnorm(n, sd=sqrt(1-buy^2))
x <- u * bux + rnorm(n, sd=sqrt(1-bux^2))


iv_bias(g, x, y)


ols_bias(x,y)
```

```{r}
param1 <- expand.grid(
    bux=seq(-1, 1, by=0.1),
    buy=seq(-1, 1, by=0.1)
)

for(i in 1:nrow(param1)) {
    y <- u * param1$buy[i] + rnorm(n, sd=sqrt(1-param1$buy[i]^2))
    x <- u * param1$bux[i] + rnorm(n, sd=sqrt(1-param1$bux[i]^2))
    param1$ols_bias[i] <- ols_bias(x, y)
    param1$iv_bias[i] <- iv_bias(g, x, y)
    param1$ols_bias_e[i] <- param1$buy[i] * param1$bux[i]
    param1$iv_bias_e[i] <- param1$buy[i] / param1$bux[i]
}

param1 %>% filter(bux != 0) %>%
gather(key="key", value="value", c(ols_bias, iv_bias)) %>% 
ggplot(., aes(x=bux, y=value)) +
geom_point(aes(colour=key)) +
facet_wrap(~ buy)
```

Is expected bias correct

```{r}
plot(ols_bias ~ ols_bias_e, param1)
```

```{r}
plot(iv_bias ~ iv_bias_e, param1 %>% filter(bux!=0))
```

now change the variances

```{r}
param2 <- expand.grid(
    bux=seq(-1, 1, by=0.1),
    buy=0.5,
    varx=c(1, 0.1, 10),
    vary=c(1, 0.1, 10)
)

for(i in 1:nrow(param2)) {
    y <- u * param2$buy[i] + rnorm(n, sd=sqrt(1-param2$buy[i]^2))
    x <- u * param2$bux[i] + rnorm(n, sd=sqrt(1-param2$bux[i]^2))
    y <- y * sqrt(param2$vary[i])
    x <- x * sqrt(param2$varx[i])
    param2$ols_bias[i] <- ols_bias(x, y)
    param2$iv_bias[i] <- iv_bias(g, x, y)
    param2$ols_bias_e[i] <- param2$buy[i] * param2$bux[i]
    param2$iv_bias_e[i] <- param2$buy[i] / param2$bux[i]
}

param2 <- param2 %>% filter(bux != 0)
param2 <- bind_rows(
    param2 %>% select(-c(ols_bias_e, iv_bias_e)) %>% gather(key="key", value="value", c(ols_bias, iv_bias)) %>% mutate(what="obs"),
    param2 %>% select(-c(ols_bias, iv_bias)) %>% gather(key="key", value="value", c(ols_bias_e, iv_bias_e)) %>% mutate(what="exp") 
)
ggplot(param2, aes(x=bux, y=value)) +
geom_point(aes(colour=key)) +
facet_grid(varx ~ vary, labeller=label_both, scale="free_y")
```

change the variance only, see if it changes test statistic

```{r}
param3 <- expand.grid(
    bux=seq(-1, 1, by=0.25),
    buy=seq(-1, 1, by=0.25),
    varx=c(0.1, 1, 10),
    vary=c(0.1, 1, 10)
) %>% mutate(sim=1:n())

param3 <- lapply(1:nrow(param3), function(i) {
    y <- u * param2$buy[i] + rnorm(n, sd=sqrt(1-param2$buy[i]^2))
    x <- u * param2$bux[i] + rnorm(n, sd=sqrt(1-param2$bux[i]^2))
    y <- y * sqrt(param2$vary[i])
    x <- x * sqrt(param2$varx[i])
    get_effs(x, y, as.matrix(g)) %>% mr %>% suppressMessages %>% mutate(sim=i)
}) %>% bind_rows() %>% inner_join(., param3, by="sim")

param3$tval <- param3$b/param3$se

param3 %>% group_by(bux, buy) %>%
    summarise(m=mean(tval), s=sd(tval))
```


```{r}
param3 %>% ggplot(., aes(x=as.factor(bux), y=abs(tval))) +
geom_point(aes(colour=as.factor(varx), shape=as.factor(vary))) +
facet_wrap(~ buy)
```


---

```{r}
sessionInfo()
```
