---
title: "Correlated SNPs"
author: "Gibran Hemani"
date: "2022-08-12"
---

## MR type 1 error rate

One instrument for X and X has no influence on Y

```{r}
library(TwoSampleMR)
library(simulateGP)
library(dplyr)
```


```{r}
set.seed(12345)
map <- tibble(snp=1, af=0.5)
params_x <- generate_gwas_params(map=map, h2=0.01, S=-0.4, Pi=1)
params_y <- generate_gwas_params(map=map, h2=0.0, S=-0.4, Pi=1)
nid <- 100000
ss <- summary_set(
    beta_gx=params_x$beta,
    beta_gy=params_y$beta,
    af=params_x$af,
    n_gx=10000,
    n_gy=10000,
    n_overlap=0,
    cor_xy=0.5
)
```

Perform MR with single causal variant

```{r}
mr(ss) %>% glimpse()
```

Perform MR with causal variant + 100 correlated tag SNPs

```{r}
ss2 <- ss[rep(1,100),] %>% mutate(SNP=1:100)
mr(ss2, method_list="mr_ivw") %>% glimpse()
```

Very small p-value - inflated type 1 error

## GRS correlation performance

Tag SNPs are perfectly correlated with causal variant

```{r}
sim <- function(nid=10000, nsnp=10)
{
  g <- matrix(0, nid, nsnp)
  g[,1] <- rnorm(nid)
  for(i in 2:nsnp)
  {
    g[,i] <- g[,1]
  }
  y <- g[,1] + rnorm(nid)
  summary(lm(y ~ g[,1]))
  grs <- rowSums(g)
  return(c(cor(y, g[,1])^2, cor(y, grs)^2))
}
sapply(1:10, function(i) sim()) %>% rowMeans() %>% tibble(method=c("Causal variant only", "GRS"), rsq=.)
```

GRS and single causal variant work the same as Jack showed.


Tag SNPs are imperfectly correlated with causal variant

```{r}
sim <- function(nid=10000, nsnp=10)
{
  g <- matrix(0, nid, nsnp)
  g[,1] <- rnorm(nid)
  for(i in 2:nsnp)
  {
    g[,i] <- g[,i] + rnorm(nid, sd=0.5)
  }
  y <- g[,1] + rnorm(nid)
  summary(lm(y ~ g[,1]))
  grs <- rowSums(g)
  return(c(cor(y, g[,1])^2, cor(y, grs)^2))
}
sapply(1:10, function(i) sim()) %>% rowMeans() %>% tibble(method=c("Causal variant only", "GRS"), rsq=.)
```

Now the GRS doesn't work well because it includes the variance of the SNP + noise that isn't causally related to the trait.

$$
r^2 = \frac{cov(grs, y)^2}{var(grs) var(y)}
$$

i.e. cov(grs,y) isn't increasing, but (var(y)) is.


## Checking

```{r}
n <- 10000
nsnp <- 10
g <- matrix(0, n, nsnp)
g[,1] <- rnorm(n)
for(i in 2:nsnp)
{
  g[,i] <- g[,i] + rnorm(n, sd=0.5)
}
y <- g[,1] + rnorm(n)
grs <- rowSums(g)
cov(y, grs)
cov(y, g[,1])
sd(grs)
sd(g[,1])
```


```{r}
sessionInfo()
```