---
title: "Fixed effects family model simulation"
author: Gibran Hemani
date: "2026-07-20"
categories: []
---


```{r}
library(dplyr)
```


## Background

We simulate two populations that differ in:

1. allele frequency at a focal genetic variant; and
2. mean environment.

Mating occurs randomly within population. Each family has two full siblings. The phenotype is generated as

$$
Y_{if} = \beta_G G_{if} + \Delta_E Z_f + \varepsilon_{if},
$$

where $Z_f=1$ denotes population 2, $\beta_G$ is the direct genetic effect, and $\Delta_E$ is the population environmental difference.

We then compare:

$$
Y \sim G + PC_1
$$

with the within-between sibling model

$$
Y_{if} \sim \bar G_f\,b_{\text{between}}
+ (G_{if}-\bar G_f)\,b_{\text{within}} + PC_1.
$$

The ancestry covariate is simulated as

$$
PC_{1f}=Z_f+\eta_f,\qquad \eta_f\sim N(0,\sigma^2_{PC}).
$$

Thus:

- `pc_sd = 0` gives perfect adjustment for population;
- larger `pc_sd` gives progressively less effective adjustment;
- the same PC value is assigned to both siblings.

This deliberately simplified PC is useful because the amount of residual demographic confounding is under direct control.

## Simulation functions

```{r}
simulate_sibling_data <- function(
    n_families = 5000,
    prop_pop2 = 0.50,
    p_pop1 = 0.20,
    p_pop2 = 0.60,
    env_pop1 = 0,
    env_pop2 = 1,
    beta_g = 1,
    noise_sd = 1,
    pc_sd = 0.75,
    seed = NULL
) {
  if (!is.null(seed)) {
    set.seed(seed)
  }

  stopifnot(
    n_families > 1,
    prop_pop2 > 0, prop_pop2 < 1,
    p_pop1 > 0, p_pop1 < 1,
    p_pop2 > 0, p_pop2 < 1,
    noise_sd >= 0,
    pc_sd >= 0
  )

  # Family population membership: 0 = population 1, 1 = population 2.
  pop2 <- rbinom(n_families, size = 1, prob = prop_pop2)
  allele_frequency <- ifelse(pop2 == 1, p_pop2, p_pop1)

  # Random mating within population. Parental genotypes are generated under HWE.
  mother_g <- rbinom(n_families, size = 2, prob = allele_frequency)
  father_g <- rbinom(n_families, size = 2, prob = allele_frequency)

  # Each sibling independently receives one randomly selected allele
  # from each parent.
  maternal_allele <- matrix(
    rbinom(
      2 * n_families,
      size = 1,
      prob = rep(mother_g / 2, times = 2)
    ),
    nrow = n_families,
    ncol = 2
  )

  paternal_allele <- matrix(
    rbinom(
      2 * n_families,
      size = 1,
      prob = rep(father_g / 2, times = 2)
    ),
    nrow = n_families,
    ncol = 2
  )

  sibling_g <- maternal_allele + paternal_allele

  fam_mean <- rowMeans(sibling_g)
  centred_g <- sweep(sibling_g, MARGIN = 1, STATS = fam_mean, FUN = "-")

  # A family-level imperfect proxy for population structure.
  pc1_family <- pop2 + rnorm(n_families, mean = 0, sd = pc_sd)

  env_family <- ifelse(pop2 == 1, env_pop2, env_pop1)

  phenotype <- beta_g * sibling_g +
    matrix(rep(env_family, times = 2), nrow = n_families, ncol = 2) +
    matrix(
      rnorm(2 * n_families, mean = 0, sd = noise_sd),
      nrow = n_families,
      ncol = 2
    )

  data.frame(
    family = rep(seq_len(n_families), times = 2),
    sibling = rep(1:2, each = n_families),
    pop2 = rep(pop2, times = 2),
    pc1 = rep(pc1_family, times = 2),
    g = as.vector(sibling_g),
    fam_mean = rep(fam_mean, times = 2),
    centred_geno = as.vector(centred_g),
    env = rep(env_family, times = 2),
    y = as.vector(phenotype)
  )
}


estimate_models <- function(dat) {
  population_model <- lm(y ~ g + pc1, data = dat)

  within_between_model <- lm(
    y ~ fam_mean + centred_geno + pc1,
    data = dat
  )

  # Residual genotype variances after adjustment for the same PC.
  g_resid <- resid(lm(g ~ pc1, data = dat))
  mean_resid <- resid(lm(fam_mean ~ pc1, data = dat))
  centred_resid <- resid(lm(centred_geno ~ pc1, data = dat))

  var_g_resid <- var(g_resid)
  var_mean_resid <- var(mean_resid)
  var_centred_resid <- var(centred_resid)

  weight_between <- var_mean_resid /
    (var_mean_resid + var_centred_resid)

  b_pop <- unname(coef(population_model)["g"])
  b_between <- unname(coef(within_between_model)["fam_mean"])
  b_within <- unname(coef(within_between_model)["centred_geno"])

  c(
    b_pop = b_pop,
    b_between = b_between,
    b_within = b_within,
    weight_between = weight_between,
    reconstructed_pop =
      weight_between * b_between +
      (1 - weight_between) * b_within,
    var_g_resid = var_g_resid,
    var_mean_resid = var_mean_resid,
    var_centred_resid = var_centred_resid
  )
}
```

## Theory

Write population membership as $Z$, with

$$
P(Z=1)=q.
$$

The expected genotype is $2p_1$ in population 1 and $2p_2$ in population 2. Define

$$
V_Z=q(1-q)
$$

and

$$
C_{GZ}=\operatorname{Cov}(G,Z)
      =2(p_2-p_1)V_Z.
$$

The average genotype variance within populations is

$$
V_{G,\text{within pop}}
=
(1-q)\,2p_1(1-p_1)
+
q\,2p_2(1-p_2).
$$

For full-sibling pairs,

$$
\operatorname{Var}(\bar G_f\mid Z)
=
\frac{3}{4}\operatorname{Var}(G\mid Z)
$$

and

$$
\operatorname{Var}(G_{if}-\bar G_f\mid Z)
=
\frac{1}{4}\operatorname{Var}(G\mid Z).
$$

The population difference in genotype means contributes only to the between-family component. Therefore,

$$
\begin{aligned}
V_G &=
V_{G,\text{within pop}}
+
4(p_2-p_1)^2V_Z,\\
V_{\bar G} &=
\frac{3}{4}V_{G,\text{within pop}}
+
4(p_2-p_1)^2V_Z,\\
V_C &=
\frac{1}{4}V_{G,\text{within pop}},
\end{aligned}
$$

where $C_{if}=G_{if}-\bar G_f$.

Because $PC_1=Z+\eta$, with
$\operatorname{Var}(\eta)=\sigma^2_{PC}$,

$$
\operatorname{Var}(PC_1)=V_Z+\sigma^2_{PC}.
$$

After linear adjustment for $PC_1$,

$$
\operatorname{Cov}(G,Z\mid PC_1)
=
C_{GZ}
\frac{\sigma^2_{PC}}{V_Z+\sigma^2_{PC}}.
$$

The residual variances are

$$
\begin{aligned}
V_{G\mid PC}
&=
V_G-\frac{C_{GZ}^2}{V_Z+\sigma^2_{PC}},\\
V_{\bar G\mid PC}
&=
V_{\bar G}-\frac{C_{GZ}^2}{V_Z+\sigma^2_{PC}},\\
V_{C\mid PC}
&=V_C.
\end{aligned}
$$

It follows that

$$
b_{\text{within}}=\beta_G,
$$

$$
b_{\text{pop}}
=
\beta_G
+
\Delta_E
\frac{\operatorname{Cov}(G,Z\mid PC_1)}
     {V_{G\mid PC}},
$$

and

$$
b_{\text{between}}
=
\beta_G
+
\Delta_E
\frac{\operatorname{Cov}(G,Z\mid PC_1)}
     {V_{\bar G\mid PC}}.
$$

Since $V_{\bar G\mid PC}<V_{G\mid PC}$, positive residual confounding gives

$$
b_{\text{between}}>b_{\text{pop}}>b_{\text{within}}.
$$

Finally,

$$
b_{\text{pop}}
=
w\,b_{\text{between}}
+
(1-w)b_{\text{within}},
$$

where

$$
w=
\frac{V_{\bar G\mid PC}}
     {V_{\bar G\mid PC}+V_{C\mid PC}}.
$$

The familiar $w=0.75$ applies after population structure has been perfectly removed. With residual population structure, population differences in allele frequency add extra variance to $\bar G_f$, so $w$ can exceed $0.75$.

```{r}
theoretical_coefficients <- function(
    prop_pop2 = 0.50,
    p_pop1 = 0.20,
    p_pop2 = 0.60,
    env_pop1 = 0,
    env_pop2 = 1,
    beta_g = 1,
    pc_sd = 0.75
) {
  q <- prop_pop2
  var_z <- q * (1 - q)

  mean_g_pop1 <- 2 * p_pop1
  mean_g_pop2 <- 2 * p_pop2
  mean_difference_g <- mean_g_pop2 - mean_g_pop1

  cov_g_z <- mean_difference_g * var_z

  mean_within_pop_var_g <-
    (1 - q) * 2 * p_pop1 * (1 - p_pop1) +
    q * 2 * p_pop2 * (1 - p_pop2)

  var_population_means <- mean_difference_g^2 * var_z

  var_g <- mean_within_pop_var_g + var_population_means
  var_fam_mean <- 0.75 * mean_within_pop_var_g +
    var_population_means
  var_centred <- 0.25 * mean_within_pop_var_g

  var_pc <- var_z + pc_sd^2

  cov_g_z_resid <- cov_g_z * pc_sd^2 / var_pc

  var_g_resid <- var_g - cov_g_z^2 / var_pc
  var_fam_mean_resid <- var_fam_mean - cov_g_z^2 / var_pc
  var_centred_resid <- var_centred

  env_difference <- env_pop2 - env_pop1

  b_within <- beta_g

  b_pop <- beta_g +
    env_difference * cov_g_z_resid / var_g_resid

  b_between <- beta_g +
    env_difference * cov_g_z_resid / var_fam_mean_resid

  weight_between <- var_fam_mean_resid /
    (var_fam_mean_resid + var_centred_resid)

  c(
    b_pop = b_pop,
    b_between = b_between,
    b_within = b_within,
    weight_between = weight_between,
    reconstructed_pop =
      weight_between * b_between +
      (1 - weight_between) * b_within,
    var_g_resid = var_g_resid,
    var_fam_mean_resid = var_fam_mean_resid,
    var_centred_resid = var_centred_resid
  )
}
```

## One large simulated sample

We use:

- causal genetic effect: $\beta_G=1$;
- allele frequencies: $p_1=0.20$, $p_2=0.60$;
- population environmental difference: $\Delta_E=1$;
- imperfect PC adjustment: `pc_sd = 0.75`.

```{r}
parameters <- list(
  prop_pop2 = 0.50,
  p_pop1 = 0.20,
  p_pop2 = 0.60,
  env_pop1 = 0,
  env_pop2 = 1,
  beta_g = 1,
  noise_sd = 1,
  pc_sd = 0.75
)

dat <- do.call(
  simulate_sibling_data,
  c(
    list(n_families = 100000, seed = 12345),
    parameters
  )
)

est <- estimate_models(dat)

theory <- do.call(
  theoretical_coefficients,
  parameters[names(parameters) != "noise_sd"]
)

comparison <- data.frame(
  quantity = c(
    "Population coefficient",
    "Between-family coefficient",
    "Within-family coefficient",
    "Between-family variance weight",
    "Weighted reconstruction of population coefficient"
  ),
  theory = unname(theory[c(
    "b_pop",
    "b_between",
    "b_within",
    "weight_between",
    "reconstructed_pop"
  )]),
  simulation = unname(est[c(
    "b_pop",
    "b_between",
    "b_within",
    "weight_between",
    "reconstructed_pop"
  )])
)

knitr::kable(comparison, digits = 4)
```

For these parameter values, the theoretical values are approximately:

```{r}
knitr::kable(
  data.frame(
    quantity = names(theory),
    value = unname(theory)
  ),
  digits = 4
)
```

The expected ordering is

$$
b_{\text{between}}\approx 1.337
>
b_{\text{pop}}\approx 1.271
>
b_{\text{within}}=1.
$$

The population coefficient is not equal to the between-family coefficient. It is a residual-variance-weighted average of the between- and within-family coefficients.

## Repeated simulation

A repeated simulation verifies both the coefficient expectations and the weighted identity.

```{r}
run_repeated_simulation <- function(
    n_sim = 200,
    n_families = 2000,
    parameters,
    first_seed = 1000
) {
  result <- t(vapply(
    seq_len(n_sim),
    FUN = function(i) {
      dat_i <- do.call(
        simulate_sibling_data,
        c(
          list(
            n_families = n_families,
            seed = first_seed + i
          ),
          parameters
        )
      )

      estimate_models(dat_i)[c(
        "b_pop",
        "b_between",
        "b_within",
        "weight_between",
        "reconstructed_pop"
      )]
    },
    FUN.VALUE = numeric(5)
  ))

  as.data.frame(result)
}

sim_results <- run_repeated_simulation(
  n_sim = 200,
  n_families = 2000,
  parameters = parameters
)

summary_table <- data.frame(
  coefficient = c(
    "Population",
    "Between family",
    "Within family"
  ),
  theory = unname(theory[c(
    "b_pop",
    "b_between",
    "b_within"
  )]),
  simulation_mean = c(
    mean(sim_results$b_pop),
    mean(sim_results$b_between),
    mean(sim_results$b_within)
  ),
  simulation_sd = c(
    sd(sim_results$b_pop),
    sd(sim_results$b_between),
    sd(sim_results$b_within)
  )
)

summary_table$mean_minus_theory <-
  summary_table$simulation_mean - summary_table$theory

knitr::kable(summary_table, digits = 4)
```

```{r}
boxplot(
  sim_results[c("b_pop", "b_between", "b_within")],
  names = c("Population", "Between", "Within"),
  ylab = "Estimated genetic coefficient",
  main = "Repeated sibling-pair simulations"
)

points(
  x = 1:3,
  y = unname(theory[c("b_pop", "b_between", "b_within")]),
  pch = 19
)

legend(
  "topright",
  legend = "Theoretical value",
  pch = 19,
  bty = "n"
)
```

The reconstruction error should be numerical sampling error:

```{r}
reconstruction_error <-
  sim_results$b_pop - sim_results$reconstructed_pop

c(
  mean_error = mean(reconstruction_error),
  sd_error = sd(reconstruction_error),
  maximum_absolute_error = max(abs(reconstruction_error))
)
```

For balanced sibling pairs and a family-level PC, the equality is essentially exact within each simulated data set because

$$
G_{if}=\bar G_f+(G_{if}-\bar G_f)
$$

and the two components are orthogonal.

## Effect of PC quality

The following calculation varies the noise in the ancestry covariate. At `pc_sd = 0`, population is measured perfectly and all three coefficients equal the causal effect. As the PC becomes less informative, residual population confounding increases.

```{r}
pc_grid <- seq(0, 3, length.out = 200)

theory_by_pc <- t(vapply(
  pc_grid,
  FUN = function(pc_value) {
    theoretical_coefficients(
      prop_pop2 = parameters$prop_pop2,
      p_pop1 = parameters$p_pop1,
      p_pop2 = parameters$p_pop2,
      env_pop1 = parameters$env_pop1,
      env_pop2 = parameters$env_pop2,
      beta_g = parameters$beta_g,
      pc_sd = pc_value
    )[c("b_pop", "b_between", "b_within")]
  },
  FUN.VALUE = numeric(3)
))

matplot(
  pc_grid,
  theory_by_pc,
  type = "l",
  lty = 1:3,
  xlab = "SD of noise in PC1",
  ylab = "Theoretical coefficient",
  main = "Residual confounding as PC quality deteriorates"
)

legend(
  "bottomright",
  legend = c("Population", "Between family", "Within family"),
  lty = 1:3,
  bty = "n"
)
```

## Interpretation

This simulation isolates a simple case in which the only source of bias is a population environmental difference correlated with allele frequency.

The results illustrate four points:

1. The within-family coefficient estimates the direct genetic effect because sibling genotype deviations are independent of population-level environment under Mendelian segregation.

2. The between-family coefficient is on the same per-allele scale as the population coefficient because `fam_mean` is a genotype average, not a genotype sum.

3. The between-family coefficient is more strongly affected by between-population confounding than the ordinary population coefficient. Under positive confounding,

   $$
   b_{\text{between}}>b_{\text{pop}}>b_{\text{within}}.
   $$

4. The population estimate is the residual-variance-weighted average

   $$
   b_{\text{pop}}
   =
   w b_{\text{between}}+
   (1-w)b_{\text{within}},
   $$

   rather than being approximately identical to $b_{\text{between}}$.

In real sibling analyses, the difference between the between- and within-family estimates can additionally reflect indirect parental genetic effects, assortative mating, shared family environment, and genotype measurement error. The present simulation intentionally excludes those mechanisms.

---

```{r}
sessionInfo()
```
