---
title: "Meta regression of instrument strength"
author: "Si Fang, Gibran Hemani"
date: "2024-05-20"
categories: []
---

```{r}
library(dplyr)
library(meta)
```

## Height on EduYears

```{r}
aivw <- read.csv("~/Downloads/MR_heightonEduYears_testing_different_pval_interval.txt") %>%
    filter(method=="Inverse variance weighted") %>%
    mutate(log10p = -log10(P_iv_max), lin = order(log10p))
aivw
```

Perform meta analysis

```{r}
m1 <- meta::metagen(b, se, data=aivw)
m1
```

Substantial heterogeneity. Perform meta regression

```{r}
mu2 <- update(m1, subgroup = aivw$log10p, tau.common = TRUE, common = FALSE)
metareg(mu2)
```

Strong relationship with instrument strength

Plot

```{r}
bubble(metareg(mu2))
```

Try using a different scale of instrument strength

```{r}
mu3 <- update(m1, subgroup = aivw$lin, tau.common = TRUE, common = FALSE)
metareg(mu3)
```

Stronger relationship with instrument strength now

Plot

```{r}
bubble(metareg(mu3))
```



## Height on CHD

```{r}
aivw <- read.csv("~/Downloads/MR_heightonCHD_testing_different_pval_interval.txt") %>%
    filter(method=="Inverse variance weighted") %>%
    mutate(log10p = -log10(P_iv_max), lin = order(log10p))
aivw
```

Perform meta analysis

```{r}
m1 <- meta::metagen(b, se, data=aivw)
m1
```

Weak heterogeneity. Perform meta regression

```{r}
mu2 <- update(m1, subgroup = aivw$log10p, tau.common = TRUE, common = FALSE)
metareg(mu2)
```

Some relationship with instrument strength

Plot

```{r}
bubble(metareg(mu2))
```


Try again using a different scale for instrument strength

```{r}
mu3 <- update(m1, subgroup = aivw$lin, tau.common = TRUE, common = FALSE)
metareg(mu3)
```

Weaker relationship with instrument strength 

Plot

```{r}
bubble(metareg(mu3))
```


## Background

---

```{r}
sessionInfo()
```
